---
title: "2. Prepare TCGA data"
# output: html_notebook
---

## Initial setup

### Load libraries
```{r}
suppressPackageStartupMessages({
  library(dplyr)
  library(limma) # vennDiagram
})
```

## Load expresssion data (downloaded form cBioPortal)

In the script below:
- tcga_mat1: Raw expression values (log2-transformed) from RNA-Seq (data_RNA_Seq_v2_expression_median.txt)
- tcga_mat2: Z-scores normalized for each gene relative to the cohort (data_RNA_Seq_v2_mRNA_median_Zscores.txt)
- tcga_mat3: Z-scores normalized across all samples (data_RNA_Seq_v2_mRNA_median_all_sample_Zscores.txt)

We use tcga_mat3 for the later analysis (see 04)

```{r}
ma_fns <- file.path(tcga_dir, c(
  "data_RNA_Seq_v2_expression_median.txt",
  "data_RNA_Seq_v2_mRNA_median_Zscores.txt",
  "data_RNA_Seq_v2_mRNA_median_all_sample_Zscores.txt"
))

tcga_mat1 <- read.table(ma_fns[1], sep = "\t", header = TRUE)
tcga_mat2 <- read.table(ma_fns[2], sep = "\t", header = TRUE)
tcga_mat3 <- read.table(ma_fns[3], sep = "\t", header = TRUE)

tcga_mat1[1:5, 1:5] # Preview of first matrix
```

### Check column names
```{r}
# Ensure the column names (samples) are consistent across all matrices
identical(colnames(tcga_mat1), colnames(tcga_mat2)) # TRUE
identical(colnames(tcga_mat1), colnames(tcga_mat3)) # TRUE
```

## Pre-processing data
### Log-transform and matrix setup

```{r}
# Log-transform the matrix
tcga_m1 <- log2(as.matrix(tcga_mat1[, -(1:2)]) + 1)
rownames(tcga_m1) <- tcga_mat1$Hugo_Symbol 

tcga_m2 <- as.matrix(tcga_mat2[, -(1:2)])
rownames(tcga_m2) <- tcga_mat2$Hugo_Symbol

tcga_m3 <- as.matrix(tcga_mat3[, -(1:2)])
rownames(tcga_m3) <- tcga_mat3$Hugo_Symbol

# Check dimensions to ensure consistency
dim(tcga_m1) # 20531 genes, 1100 samples
dim(tcga_m2) # 20440 genes, 1100 samples
dim(tcga_m3) # 20531 genes, 1100 samples
```

### Check gene consistency between mat1 and mat3
```{r}
identical(tcga_mat1$Hugo_Symbol, tcga_mat3$Hugo_Symbol) # Should return TRUE
```
### Remove metastasized samples
```{r}
is_dup <- grepl("06$", colnames(tcga_m3)) # Identify metastasized samples
tcga_m1 <- tcga_m1[, !is_dup]
tcga_m2 <- tcga_m2[, !is_dup]
tcga_m3 <- tcga_m3[, !is_dup]

colnames(tcga_m1) <- colnames(tcga_m2) <- colnames(tcga_m3) <- sub("\\.01$", "", colnames(tcga_m3))
```


## Clinical data preparation
### Load clinical data and match it to expression data
```{r}
cl_path <- file.path(tcga_dir, "data_bcr_clinical_data_patient.txt")
cl_data <- read.delim(cl_path, header = TRUE, skip = 4, stringsAsFactors = FALSE)
cl_data$PATIENT_ID <- gsub("-", ".", cl_data$PATIENT_ID)

used_ids <- colnames(tcga_m3)
tcga.clinical <- filter(cl_data, PATIENT_ID %in% used_ids) %>%
	filter(ER_STATUS_BY_IHC %in% c("Positive","Negative")) %>%
	filter(PR_STATUS_BY_IHC %in% c("Positive","Negative")) %>%
	filter(IHC_HER2 %in% c("Positive","Negative"))

tcga.ids <- list(
  hr = tcga.clinical %>% filter((ER_STATUS_BY_IHC == "Positive" | PR_STATUS_BY_IHC == "Positive") & IHC_HER2 == "Negative") %>% pull(PATIENT_ID),
  her2 = tcga.clinical %>% filter(ER_STATUS_BY_IHC == "Negative" & PR_STATUS_BY_IHC == "Negative" & IHC_HER2 == "Positive") %>% pull(PATIENT_ID),
  dp = tcga.clinical %>% filter((ER_STATUS_BY_IHC == "Positive" | PR_STATUS_BY_IHC == "Positive") & IHC_HER2 == "Positive") %>% pull(PATIENT_ID),
  tnbc = tcga.clinical %>% filter(ER_STATUS_BY_IHC == "Negative" & PR_STATUS_BY_IHC == "Negative" & IHC_HER2 == "Negative") %>% pull(PATIENT_ID)
)
```

### Remove redundant genes
```{r}
used.ids <- tcga.clinical$PATIENT_ID

# Process tcga_m1
rm1 <- rowMeans(tcga_m1, na.rm = TRUE)
idx1 <- tapply(seq_along(rm1), names(rm1), function(i) i[which.max(rm1[i])])
mm1 <- tcga_m1[idx1, used.ids]
rownames(mm1) <- names(rm1[idx1])

# Process tcga_m3 using idx1 from tcga_m1
mm3 <- tcga_m3[idx1, used.ids]
rownames(mm3) <- rownames(mm1)

tcga.m1 <- mm1
tcga.m3 <- mm3

```

### Process expression data for analysis
```{r}
tcga.syms <- intersect(rownames(tcga.m1), rownames(tcga.m3))
tcga.syms <- tcga.syms[!is.na(tcga.syms)]

tcga.m1 <- tcga.m1[tcga.syms, ]
tcga.m3 <- tcga.m3[tcga.syms, ]

tcga.m1 <- tcga.m1[, used.ids]
tcga.m3 <- tcga.m3[, used.ids]
```

## Venn diagram for subtypes 
```{r}
mm <- with(tcga.clinical, cbind(
  `ER+` = ER_STATUS_BY_IHC == "Positive",
  `PR+` = PR_STATUS_BY_IHC == "Positive",
  `HER2+` = IHC_HER2 == "Positive"
))

mm[] <- sapply(mm, as.integer)
rownames(mm) <- tcga.clinical$PATIENT_ID

pdf(file.path(fig_dir,"02_TCGA_prep","tcga_brca_subtypes.pdf"),width=7,height=7)
vennDiagram(mm,main="TCGA samples, by subtypes")
dev.off()

```

## Save processed data
```{r}
# Save the relevant objects to a .rda file for easy loading in future analysis
save(tcga.ids,tcga.clinical,tcga.syms,tcga.m1,tcga.m3,
     file=file.path(obj_dir,"TCGA_META","tcga_processed.rda"))
```
