---
title: "4. gene correlations in TCGA and METABRIC"
# output: html_notebook
---

## Initial setup

### Load library
```{r}
suppressPackageStartupMessages({
  library(RColorBrewer)
  library(dplyr)
  library(heatmap3)
  library(limma)
  library(org.Hs.eg.db)
})
```

### Load data
```{r}
x <- load(file=file.path(obj_dir,"Gene_modules","gene_signature.rda")) # gene.sig
y <- load(file=file.path(obj_dir,"TCGA_META","tcga_processed.rda")) # tcga.ids,tcga.clinical,tcga.syms,tcga.m1,tcga.m3
z <- load(file=file.path(obj_dir,"TCGA_META","metabric_processed.rda")) #
#meta.ids,meta.clinical,meta.syms,meta.m1,meta.m3,meta.exp.th,meta.df,meta.sig.df 
```

#### Gene and signature setup
```{r}
ers.genes <- gene.sig$ERS
immune.genes <- unique(unlist(gene.sig[-1]))

sig.gns <- unique(unlist(gene.sig)) ## 988 genes in either of the three signatures
gene.list.gns <- unique(intersect(unlist(gene.sig),intersect(tcga.syms,meta.syms))) ## 988 -> 778 unique
tested.gns  <- intersect(tcga.syms,meta.syms) # 15397

hallmarks <- c("ESR1","CD8A","CD8B","HLA-A")
```


```{r}
saveRDS(tested.gns,file=file.path(obj_dir,"TCGA_META","common_tested_gns.rds"))
```

### TCGA correlation for `gene.list.gns`
```{r}
tcga.exp <- tcga.m3[gene.list.gns,]
tcga.sp <- list()

for(cl in names(tcga.ids)){
	used.ids <- tcga.ids[[cl]]
	tcga.exp.sub <- tcga.exp[, used.ids]
	tcga.sp[[cl]] <- cor(t(tcga.exp.sub), method="spearman",use="everything")
}
```

### METABRIC correlation for `gene.list.gns`
```{r}
meta.exp <- meta.m3[gene.list.gns,]
meta.sp <- list()

for(cl in names(meta.ids)){
	used.ids <- meta.ids[[cl]]
	meta.exp.sub <- meta.exp[, used.ids]
	tmp <- cor(t(meta.exp.sub), method="spearman",use="everything")
  
  # Fix rows/columns with missing values
  idx <- which(rowSums(is.na(tmp)) == nrow(tmp) - 1)
  tmp1 <- cor(t(meta.exp.sub[idx, ]), t(meta.exp.sub), method="spearman", use="pairwise.complete.obs")
  tmp[idx, ] <- tmp1
  tmp[, idx] <- t(tmp1)
  
  meta.sp[[cl]] <- tmp
}
```


### Save or load correlations
```{r}
if(0){
  save(tcga.sp,meta.sp,file=file.path(obj_dir,"TCGA_META","tcga_metabric_cors.rda"))
}else{
  load(file=file.path(obj_dir,"TCGA_META","tcga_metabric_cors.rda")) # tcga.sp, meta.sp
}
```

## QC: transcriptome data

### [Fig S1B] Correlation between hallmark genes

```{r}
source(file.path(src_dir,"utils","tri.heatmap.R"))
```

```{r}
tcga.sp1 <- tcga.sp$hr
meta.sp1 <- meta.sp$hr

tcga.hm <- tcga.sp1[hallmarks,hallmarks]
tcga.hm[!upper.tri(tcga.hm)] <- NA
meta.hm <- meta.sp1[hallmarks,hallmarks]
meta.hm[!upper.tri(meta.hm)] <- NA
```


```{r}
pdf(file.path(fig_dir,"04_TCGA_METABRIC_module_identification","corr_hallmark_tcga_ext.pdf"),width=5,height=5)
tri.heatmap(tcga.hm,main="TCGA")
dev.off()

pdf(file.path(fig_dir,"04_TCGA_METABRIC_module_identification","corr_hallmark_meta_ext.pdf"),width=5,height=5)
tri.heatmap(meta.hm,main="METABRIC")
dev.off()
```

### Compare mean expression in TCGA and METABRIC
```{r}
tcga.m1 <- tcga.m1[tested.gns,]
meta.m1 <- meta.m1[tested.gns,]
tcga.m3 <- tcga.m3[tested.gns,]
meta.m3 <- meta.m3[tested.gns,]

identical(rownames(tcga.m1),rownames(meta.m1))

med.t <- rowMeans(tcga.m1,na.rm=T)
med.m <- rowMeans(meta.m1,na.rm=T)

med.ht <- rowMeans(tcga.m1[gene.list.gns,tcga.ids$hr],na.rm=T)
med.hm <- rowMeans(meta.m1[gene.list.gns,meta.ids$hr],na.rm=T)
cor.t <- tcga.sp$hr[gene.list.gns,gene.list.gns]
cor.m <- meta.sp$hr[gene.list.gns,gene.list.gns]
diag(cor.t) <- NA
diag(cor.m) <- NA
med.cor.t <- rowMeans(abs(cor.t),na.rm=T)
med.cor.m <- rowMeans(abs(cor.m),na.rm=T)
med.cors <- list(TCGA=med.cor.t,METABRIC=med.cor.m)
```

### [Fig S1CDE] Visualizations - Correlation comparisons

```{r}
## Fig S1C
pdf(file.path(fig_dir,"04_TCGA_METABRIC_module_identification","mean_expression_2cohorts.pdf"),width=7,height=7)
par(mar=c(5,5,3,3))
cols <- densCols(med.t,med.m,nbin=256)
plot(med.t,med.m,col=cols,pch=20,cex=.4,
     xlab="Mean log2 expression in TCGA",
     ylab="Mean log2 expression in METABRIC",
     main="Comparison of mean expression for individual genes")
lo <- lowess(med.t,med.m,f=1/3)
lines(lo,col=2,lwd=2)
abline(h=meta.exp.th,lty=2,col=1,lwd=2)
text(15,meta.exp.th,"Background\nbelow this line",pos=3,cex=1)
dev.off()

## Fig S1D
pdf(file.path(fig_dir,"04_TCGA_METABRIC_module_identification","mean_corr_2cohorts.pdf"),width=5,height=7)
p.val <- signif(t.test(med.cors$TCGA,med.cors$METABRIC,alterantive="two.sided",paired=T)$p.value,2)
boxplot(med.cors,ylab="Mean absolute spearman corr.",
        main=paste0("Magnitude of correlation within two modules\n(p-value:",signif(p.val,2),")"))
dev.off()

## Fig S1E_left
rng <- range(med.cors)
pdf(file.path(fig_dir,"04_TCGA_METABRIC_module_identification","mean_exp_corr_tcga.pdf"),width=7,height=7)
par(mar=c(5,5,3,3))
cols <- densCols(med.ht,med.cor.t,nbin=256)
plot(med.ht,med.cor.t,col=cols,pch=20,cex=1,ylim=rng,
     xlab="Mean log2 expression in TCGA",
     ylab="Mean absolute correlation in TCGA",
     main="TCGA")
dev.off()

## Fig S1E_right
pdf(file.path(fig_dir,"04_TCGA_METABRIC_module_identification","mean_exp_corr_metabric.pdf"),width=7,height=7)
par(mar=c(5,5,3,3))
cols <- densCols(med.hm,med.cor.m,nbin=256)
plot(med.hm,med.cor.m,col=cols,pch=20,cex=1,ylim=rng,
     xlab="Mean log2 expression in METABRIC",
     ylab="Mean absolute correlation in METABRIC",
     main="METABRIC")
abline(v=meta.exp.th,lty=2,col=1,lwd=2)
text(meta.exp.th,0.27,"Background\nleft to this line",pos=4,cex=1)
dev.off()
```

## Identify modules


### Heatmap and dendrogram - strating point

### Gene selection in the two cohorts

```{r}
tcga.sp1 <- tcga.sp$hr
meta.sp1 <- meta.sp$hr

# cols <- colorRampPalette(rev(brewer.pal(11,"RdBu")))(100)
cols <- colorRampPalette(c("navy", "white", "firebrick3"))(1024)


hc.t <- heatmap3(tcga.sp1[gene.list.gns,gene.list.gns],
          col=cols,distfun=dist,scale="none",
          balanceColor=TRUE,
          main="TCGA (778 genes)")
hc.m <- heatmap3(meta.sp1[gene.list.gns,gene.list.gns],
          col=cols,distfun=dist,scale="none",
          balanceColor=TRUE,
          main="METABRIC (778 genes)")
```

### Filter co-expressing genes
#### Gene selection in TCGA
```{r}
hcr.t1 <- hcr.t <- hc.t$hcr
hcr.t1$labels[!hcr.t1$labels %in% hallmarks] <- ""

if(0){
  ## define the thers.thold for cutting the dendrogram
  plot(hcr.t1,lwd=.5,cex=.3)
  abline(h=6,col=3)
  abline(h=12.9,col=3)
}
## first define the immune and proliferation
c1.t <- cutree(hcr.t,h=12.9)
cl1.t <- tapply(names(c1.t),c1.t,identity)

cc.t <- intersect(cl1.t[[5]],gene.sig$ERS) # proliferation
immune.t <- intersect(cl1.t[[1]],c(gene.sig$APM,gene.sig$TC)) # immune

## first define the immune and proliferation
c2 <- cutree(hcr.t,h=6)
cl2 <- tapply(names(c2),c2,identity)
idx <- grep("ESR1",cl2)

ers.t <- intersect(cl2[[idx]],gene.sig$ERS)

tcga.sig <- list(ERS=ers.t,CC=cc.t,Immune=immune.t)
```

#### Gene selection in METABRIC
```{r}
hcr.m1 <- hcr.m <- hc.m$hcr
hcr.m1$labels[!hcr.m1$labels %in% hallmarks] <- ""

if(0){
  plot(hcr.m1,lwd=.5,cex=.3)
  abline(h=8,col=3)
  abline(h=13,col=3)
}

## first define the immune and proliferation
c1.m <- cutree(hcr.m,h=13)
cl1.m <- tapply(names(c1.m),c1.m,identity)

cc.m <- intersect(cl1.m[[3]],gene.sig$ERS) # proliferation
immune.m <- intersect(cl1.m[[1]],c(gene.sig$APM,gene.sig$TC)) # immune

## first define the immune and proliferation
c2 <- cutree(hcr.m,h=8)
cl2 <- tapply(names(c2),c2,identity)
idx <- c2["ESR1"]

ers.m <- intersect(cl2[[idx]],gene.sig$ERS)

meta.sig <- list(ERS=ers.m,CC=cc.m,Immune=immune.m)
```

#### Combine both signatures - select genes that are only positively correated within the module

```{r}
nm.sigs <- c("ERS","CC","Immune")

sigs <- list()
for(sig in nm.sigs){
  tmp1 <- tcga.sig[[sig]]
  tmp2 <- meta.sig[[sig]]
  tmp12 <- tmp1[tmp1 %in% tmp2]
  sigs[[sig]] <- tmp12
}

if(0){
  ## ERS module
  sp1 <- tcga.sp1[sigs$ERS,sigs$ERS] 
  diag(sp1) <- NA
  hist(sp1,breaks=100)
  sum(sp1 < 0,na.rm=T) # ERS module is all positively correlated
  
  ## APM/TC module
  sp2 <- tcga.sp1[sigs$Immune,sigs$Immune]
  diag(sp2) <- NA
  hist(sp2,breaks=100)
  sum(sp2 < 0,na.rm=T) # 1992 combination is negative
  
  g2 <- colSums(sp2 < 0,na.rm=T)
  m2 <- colMeans(sp2,na.rm=T)
  any(m2 < 0) # FALSE - none of the correlation is negative on average
  plot(sort(g2,decreasing=T))
}

## trim positive cor between ers and immune
sp12 <- tcga.sp1[sigs$ERS,sigs$Immune] 
hist(sp12,breaks=100)
g12 <- colSums(sp12 > 0,na.rm=T)
m12 <- colMeans(sp12)
any(m12 > 0) ## 6 APM/TC genes are positively correlated with ERS on average

pos <- names(which(m12 >= 0)) ## the six genes to be removed
sigs$Immune <- sigs$Immune[!sigs$Immune %in% pos] # refine sigs-immune

##
g12 <- rowSums(sp12 > 0,na.rm=T)/ncol(sp12)
m12 <- rowMeans(sp12)
any(m12 > 0)

##
if(0){
  plot(g12,m12,main="mean correlation of ERS genes",
       xlab="ratio of ERS genes whose correlation against APM/TC are positive",
       ylab="mean correlation against APM/TC")
  abline(h=0,v=0)

  plot(g2,m2)
  abline(h=0,v=0)
  
  hist(sp1,breaks=100)
  hist(sp2,breaks=100)
}
```
### Subdividing the APM/TC module
```{r}
module.gns <- c(sigs$ERS,sigs$Immune)
a <- heatmap3(tcga.sp1[module.gns,module.gns],col=cols,balanceColor=TRUE,scale="none")

##
hcr1 <- hcr <- a$hcr
hcr1$labels[!hcr1$labels %in% hallmarks] <- ""
if(0){
  plot(hcr1,lwd=.5,cex=.3)
  abline(h=1,col=3)
}

c3.0 <- cutree(hcr,h=1)
c3 <- as.numeric(factor(c3.0,levels=c(1,2,4,3)))
names(c3) <- names(c3.0)

mod.gns <- tapply(names(c3),c3,identity)
names(mod.gns) <- c("ERS","TNFa/NFkB","IFN-I","APM/TC")
```

#### summary - size of the gene modules
```{r}
sapply(sigs,length)
sapply(mod.gns,length)
```

## Summary gene selection

### [side_colors] save side colors for heatmap
```{r}
lcol1 <- rep("white",length(gene.list.gns))
names(lcol1) <- gene.list.gns
lcol1[c("ESR1","CD8A","HLA-A")] <- rep(c("steelblue","indianred"),c(1,2))

## lcol2: initial gene sets
gs1 <- sapply(gene.sig,function(gs)gene.list.gns %in% gs)
gs2 <- cbind(ERS=gs1[,1],Immune=(gs1[,2]|gs1[,3]))
gs.type <- factor(gs2 %*% c(1,2))
levels(gs.type) <- c("ERS","Immune","Both")

lcol2 <- c(brewer.pal(3,"Pastel1")[2:1],"white")[as.numeric(gs.type)]

# lcol3: final clusters
lcol3 <- rep("white",length(gene.list.gns))
names(lcol3) <- gene.list.gns
lcol3[sigs$ERS] <- "steelblue"
lcol3[sigs$Immune] <- "indianred"
lcol3[sigs$CC] <- "grey"

## lcol4: APM/TC subclasses
lcol4 <- rep("white",length(gene.list.gns))
names(lcol4) <- gene.list.gns
lcol4[names(c3)] <- c("white",brewer.pal(3,"Accent")[c(1,3,2)])[c3]

gl.side.cols <- cbind(hallmark=lcol1,`initial collection`=lcol2,`module`=lcol3,`immune`=lcol4)
rownames(gl.side.cols) <- gene.list.gns
mod.side.cols <- gl.side.cols[module.gns,-2]

slcol3 <- rep(c("steelblue","indianred"),c(1,3))
slcol4 <- c("white",brewer.pal(3,"Accent")[c(1,3,2)])

uniq.mod.side.cols <- cbind(`module`=slcol3,`immune`=slcol4)
rownames(uniq.mod.side.cols) <- names(mod.gns)
```

```{r}
if(0){
  save(gl.side.cols,mod.side.cols,uniq.mod.side.cols,file=file.path(obj_dir,"misc","tcga_sidecols.rda"))
  # lcols -> gl.side.cols
  # slcols -> uniq.m.side.cols
}
```


### [Fig 1B, S2A] Heatmaps - all 778 genes
```{r}
tcga.sp3 <- tcga.sp1
meta.sp3 <- meta.sp1
is.hmk <- gene.list.gns %in% hallmarks
rownames(tcga.sp3)[!is.hmk] <- colnames(tcga.sp3)[!is.hmk] <- NA
rownames(meta.sp3)[!is.hmk] <- colnames(meta.sp3)[!is.hmk] <- NA
if(1){
  pdf(file.path(fig_dir,"04_TCGA_METABRIC_module_identification","heatmap_corr_tcga_all_v2.pdf"),
      width=15,height=15)
  hc.t <- heatmap3(tcga.sp3,
            col=cols,distfun=dist,scale="none",
            balanceColor=TRUE,
            ColSideColors=gl.side.cols[,1:3],
            RowSideColors=gl.side.cols[,1:3],
            margins=c(9,9),
            cexRow=3,cexCol=3,
            main="TCGA (778 genes)")
  dev.off()
  
  pdf(file.path(fig_dir,"04_TCGA_METABRIC_module_identification","heatmap_corr_meta_all_v2.pdf"),
      width=15,height=15)
  hc.m <- heatmap3(meta.sp3,
            col=cols,distfun=dist,scale="none",
            balanceColor=TRUE,
            margins=c(9,9),            
            cexRow=3,cexCol=3,
            ColSideColors=gl.side.cols[,1:3],
            RowSideColors=gl.side.cols[,1:3],          
            main="METABRIC (778 genes)")
  dev.off()
}
```

### [Fig. 1X] mean correlation coefficients

```{r}
tcga.sp2 <- tcga.sp1
meta.sp2 <- meta.sp1

diag(tcga.sp2) <- NA
diag(meta.sp2) <- NA

mean.tcga.sp <- sapply(mod.gns,function(x){
  sapply(mod.gns,function(y){
    mean(tcga.sp2[x,y],na.rm=T)
  })
})

mean.meta.sp <- sapply(mod.gns,function(x){
  sapply(mod.gns,function(y){
    mean(meta.sp2[x,y],na.rm=T)
  })
})
```

```{r}
idx <- 1:4

# cols <- colorRampPalette(rev(brewer.pal(11,"RdBu")))(201)
cols <- colorRampPalette(c("navy", "white", "firebrick3"))(201)

n.max.cor.1 <- round(max(abs(mean.tcga.sp))*100)
cols1 <- cols[(-100:100) %in% seq(-n.max.cor.1,n.max.cor.1)]

pdf(file.path(fig_dir,"04_TCGA_METABRIC_module_identification","heatmap_mean_tcga_mods_v2.pdf"),width=7,height=7)
hc.t <- heatmap3(mean.tcga.sp[idx,idx],Rowv=NA,Colv=NA,
    # dendrogram="none",
    col=cols1,scale="none",distfun=dist,
    ColSideColors=uniq.mod.side.cols[idx,2:1],
    RowSideColors=uniq.mod.side.cols[idx,],          
    balanceColor = TRUE,
    margins=c(7,7),
    cexRow=1.5,cexCol=1.5,
    main="TCGA (mean correlation)")
dev.off()

n.max.cor.2 <- round(max(abs(mean.meta.sp))*100)
cols2 <- cols[(-100:100) %in% seq(-n.max.cor.2,n.max.cor.2)]

pdf(file.path(fig_dir,"04_TCGA_METABRIC_module_identification","heatmap_mean_meta_mods.pdf"),width=7,height=7)
hc.m <- heatmap3(mean.meta.sp[idx,idx],Rowv=NA,Colv=NA,
          # dendrogram="none",
          col=cols2,scale="none",distfun=dist,
          ColSideColors=uniq.mod.side.cols[idx,2:1],
          RowSideColors=uniq.mod.side.cols[idx,],          
          balanceColor = TRUE,          
          margins=c(7,7),       
          cexRow=1.5,cexCol=1.5,  
          main="METABRIC (mean correlation)")
dev.off()
```

### [Fig 1D, S2B] subset of genes

```{r}
tcga.sp1 <- tcga.sp$hr[module.gns,module.gns]
meta.sp1 <- meta.sp$hr[module.gns,module.gns]

d.tcga <- as.dist(1-tcga.sp1)
d.meta <-as.dist(1-meta.sp1)
dend1 <- as.dendrogram(hclust(d.tcga,method="average"))
dend2 <- as.dendrogram(hclust(d.meta,method="average"))

a <- heatmap3(tcga.sp1,
              Colv=d.tcga,Rowv=d.tcga,
              # ColSideColors=gl.side.cols[module.gns,],
              col=cols,
              scale="none",
              balanceColor=TRUE)

hcr1 <- hcr <- a$hcr
hcr1$labels[!hcr1$labels %in% hallmarks] <- ""
if(0){
  plot(hcr1,lwd=.5,cex=.3)
  abline(h=1,col=3)
}

c3 <- cutree(hcr,h=1)
cl3 <- tapply(names(c3),c3,identity)
names(cl3) <- c("ERS",paste0("Immune",1:3))

## lcol1: hallmark genes
lcol1 <- rep("white",length(module.gns))
names(lcol1) <- module.gns
lcol1[c("ESR1","CD8A","HLA-A")] <- rep(c("blue","red"),c(1,2))

# lcol3: final clusters
lcol3 <- rep("white",length(module.gns))
names(lcol3) <- module.gns
lcol3[sigs$ERS] <- "steelblue"
lcol3[sigs$Immune] <- "indianred"

## APM/TC subclasses
lcol4 <- cbind(subclass=c("white",brewer.pal(3,"Accent"))[c3])

mod.side.cols <- cbind(hallmark=lcol1,module=lcol3,subclass=lcol4)

is.hmk <- module.gns %in% hallmarks
rownames(tcga.sp1)[!is.hmk] <- colnames(tcga.sp1)[!is.hmk] <- NA
rownames(meta.sp1)[!is.hmk] <- colnames(meta.sp1)[!is.hmk] <- NA

##
rng <- range(tcga.sp1)#[module.gns,module.gns])
n.rng <- max(abs(round((rng/1)*50)))*c(-1,1)
# cols <- colorRampPalette(rev(brewer.pal(11,"RdBu")))(101)
cols <- colorRampPalette(c("navy", "white", "firebrick3"))(101)

names(cols) <- -50:50
cols <- cols[as.character(seq(n.rng[1],n.rng[2]))]

##
pdf(file.path(fig_dir,"04_TCGA_METABRIC_module_identification","heatmap_corr_tcga_sub_v2.pdf"),
    width=15,height=15)
a <- heatmap3(tcga.sp1,
              Colv=d.tcga,Rowv=d.tcga,
              col=cols,scale="none",
              ColSideColors=mod.side.cols,
              RowSideColors=mod.side.cols,
              margins=c(9,9),
              cexRow=3,cexCol=3,              
              balanceColor=TRUE)
dev.off()

rng <- range(meta.sp1)
n.rng <- max(abs(round((rng/1)*50)))*c(-1,1)
# cols <- colorRampPalette(rev(brewer.pal(11,"RdBu")))(101)
cols <- colorRampPalette(c("navy", "white", "firebrick3"))(101)

names(cols) <- -50:50
cols <- cols[as.character(seq(n.rng[1],n.rng[2]))]

pdf(file.path(fig_dir,"04_TCGA_METABRIC_module_identification","heatmap_corr_meta_sub_v2.pdf"),width=15,height=15)
b <- heatmap3(meta.sp1,
              Colv=d.meta,Rowv=d.meta,
              col=cols,scale="none",
              ColSideColors=mod.side.cols,
              RowSideColors=mod.side.cols,
              margins=c(9,9),
              cexRow=3,cexCol=3,              
              balanceColor=TRUE)
dev.off()
```

## Save data
### as csv files
```{r}
x <- list(
  ERS=data.frame(geneid=unlist(mget(sort(mod.gns$ERS),org.Hs.egSYMBOL2EG,ifnotfound=NA)),symbol=sort(mod.gns$ERS)),
  `TNFa_NFkB`=data.frame(geneid=sapply(mget(sort(mod.gns[["TNFa/NFkB"]]),org.Hs.egSYMBOL2EG,ifnotfound=NA),paste,collapse=";"),
                    symbol=sort(mod.gns[["TNFa/NFkB"]])),
  `IFN-I`=data.frame(geneid=sapply(mget(sort(mod.gns[["IFN-I"]]),org.Hs.egSYMBOL2EG,ifnotfound=NA),paste,collapse=";"),
                    symbol=sort(mod.gns[["IFN-I"]])),  
  `APM_TC`=data.frame(geneid=sapply(mget(sort(mod.gns[["APM/TC"]]),org.Hs.egSYMBOL2EG,ifnotfound=NA),paste,collapse=";"),
                    symbol=sort(mod.gns[["APM/TC"]])),
  `Cell cycle`=data.frame(geneid=sapply(mget(sort(sigs$CC),org.Hs.egSYMBOL2EG,ifnotfound=NA),paste,collapse=";"),
                    symbol=sort(sigs$CC)))
wb <- openxlsx::buildWorkbook(x)

openxlsx::saveWorkbook(wb,file.path(fig_dir,"04_TCGA_METABRIC_module_identification",
                                    "Table S2 gene_modules.xlsx"))
```
### as R objects

```{r}
saveRDS(sigs, file=file.path(obj_dir,"Gene_modules","modules3_ers_immune_cc.rds")) # ers, immune, cc
saveRDS(mod.gns, file=file.path(obj_dir,"Gene_modules","module_genes_final.rds")) # module genes
save(gl.side.cols, mod.side.cols, uniq.mod.side.cols,file=file.path(obj_dir,"misc","tcga_heatmap_sidecols.rda"))
```

