---
title: "5. Looking into the overlap between the clusters"
# output: html_notebook
---

## Initial setup

### Load libraries
```{r}
suppressPackageStartupMessages({
  library(qvalue)
  library(org.Hs.eg.db)
  library(dplyr)
})
```

### Load data
```{r}
coarse.mods <- readRDS(file=file.path(obj_dir,"1-4_prep","modules3_ers_immune_cc.rds"))
final.mods <- readRDS(file.path(obj_dir,"1-4_prep","module_genes_final_122923.rds"))
final.mods <- c(final.mods, coarse.mods[2:3])

tested.gns <- readRDS(file=file.path(obj_dir,"1-4_prep","common_tested_gns_v3.rds"))

sapply(final.mods,length)
```

## Pathway enrichment analysis
### Load pathways from msigdb v6.1
```{r}
if(0){
  gmts <- dir(path=msigdb_dir)
  
  h0 <- strsplit(readLines(file.path(msigdb_dir,"h.all.v2022.1.Hs.symbols.gmt")),"\\t")
  #h0 <- strsplit(readLines(file.path(msigdb_dir,"h.all.v2022.1.Hs.entrez.gmt")),"\\t")
  gs0 <- lapply(h0,function(x)x[-(1:2)])
  names(gs0) <- sapply(h0,function(x)x[1])
  
  c2 <- strsplit(readLines(file.path(msigdb_dir,"c2.all.v2022.1.Hs.symbols.gmt")),"\\t")
  #c2 <- strsplit(readLines(file.path(msigdb_dir,"c2.all.v2022.1.Hs.entrez.gmt")),"\\t")
  gs2 <- lapply(c2,function(x)x[-(1:2)])
  names(gs2) <- sapply(c2,function(x)x[1])
  
  c5 <- strsplit(readLines(file.path(msigdb_dir,"c5.all.v2022.1.Hs.symbols.gmt")),"\\t")
  #c5 <- strsplit(readLines(file.path(msigdb_dir,"c5.go.v2022.1.Hs.entrez.gmt")),"\\t")
  gs5 <- lapply(c5,function(x)x[-(1:2)])
  names(gs5) <- sapply(c5,function(x)x[1])
  
  msigdb <- c(gs0,gs2,gs5)
  ans <- sub("_.+","",names(msigdb))
  msigdb <- msigdb[ans %in% c("HALLMARK","GOBP","GOMF","GOCC","KEGG","REACTOME","BIOCARTA")]
  
  saveRDS(msigdb,file.path(obj_dir,"5_pathway_analysis","msigdb.rds"))
  
  reg0 <- strsplit(readLines(file.path(msigdb_dir,"c3.all.v2023.1.Hs.symbols.gmt")),"\\t")
  msigdb.reg0 <- lapply(reg0,function(x)x[-(1:2)])
  names(msigdb.reg0) <- sapply(reg0,function(x)x[1])
  
  saveRDS(msigdb.reg0,file.path(obj_dir,"5_pathway_analysis","msigdb_reg.rds"))
  
}else{
  msigdb <- readRDS(file.path(obj_dir,"5_pathway_analysis","msigdb.rds"))
}
```

#### Filter pathways (each pathway should have >= 20 and <= 500 genes)
```{r}
ms.gns <- unique(unlist(msigdb))

all.genes <- intersect(ms.gns,tested.gns) # 13824
msigdb.1 <- lapply(msigdb,function(x){
	gs1 <- x[x %in% all.genes]
	return(gs1)
})

ge20 <- sapply(msigdb.1,length)>=20 & sapply(msigdb.1,length) <= 500
msigdb.2 <- msigdb.1[ge20] # 5237 pathways
ans <- sub("_.+","",names(msigdb.2))
table(ans)

all.genes <- intersect(unlist(msigdb.2),tested.gns) # 13322

comm.gns <- lapply(final.mods,function(x){
  gs1 <- x[x %in% all.genes]
  return(gs1)
})

names(comm.gns) <- sub("/","",names(comm.gns))
```
```{r}
## keep only HALLMARKs
msigdb.2 <- msigdb.2[grepl("^HALLMARK",names(msigdb.2))]

```

### Compute significance of overlaps
```{r}
compute_pvals <- function(module, gene_set) {
  tab <- table(all.genes %in% module, all.genes %in% gene_set)[c("TRUE", "FALSE"), c("TRUE", "FALSE")]
  fisher.test(tab, alternative = "greater")$p.value
}

pvals.ERS <- sapply(msigdb.2, function(gene_set) compute_pvals(comm.gns$ERS, gene_set))
pvals.CC <- sapply(msigdb.2, function(gene_set) compute_pvals(comm.gns$CC, gene_set))
pvals.Imm <- sapply(msigdb.2, function(gene_set) compute_pvals(comm.gns$Immune, gene_set))
pvals.TNFaNFkB <- sapply(msigdb.2, function(gene_set) compute_pvals(comm.gns$TNFaNFkB, gene_set))
pvals.IFNI <- sapply(msigdb.2, function(gene_set) compute_pvals(comm.gns$`IFN-I`, gene_set))
pvals.APMTC <- sapply(msigdb.2, function(gene_set) compute_pvals(comm.gns$APMTC, gene_set))

```

### FDR-adjusted p-values
```{r}
nlqs.ERS <- -log10(qvalue(pvals.ERS)$qvalue)
nlqs.CC <- -log10(qvalue(pvals.CC)$qvalue)
nlqs.Imm <- -log10(qvalue(pvals.Imm)$qvalue)
nlqs.TNFaNFkB <- -log10(qvalue(pvals.TNFaNFkB)$qvalue)
nlqs.IFNI <- -log10(qvalue(pvals.IFNI)$qvalue)
nlqs.APMTC <- -log10(qvalue(pvals.APMTC)$qvalue)

```

#### significant pathways
```{r}
df <- data.frame(pathway=names(nlqs.ERS),
  ERS=nlqs.ERS,
  CC=nlqs.CC,
  Imm=nlqs.Imm,
  TNFaNFkB=nlqs.TNFaNFkB,
  IFNI=nlqs.IFNI,
  APMTC=nlqs.APMTC) %>%
  
  mutate(across(ERS:APMTC, ~ round(.,2)))

rownames(df) <- c() 

max.class1 <- names(df[2:4])[apply(df[2:4],1,which.max)]
max.class2 <- names(df[5:7])[apply(df[5:7],1,which.max)]
max.nlq <- apply(df[c(2:3,5:7)],1,max)

df1 <- df %>%
  mutate(class1 = factor(max.class1,levels=names(df)[2:4])) %>%
  mutate(class2 = factor(max.class2,levels=names(df)[5:7])) %>%  
  mutate(max.nlq = max.nlq) %>%
  arrange(class1,class2,desc(max.nlq)) %>%
  filter(max.nlq > 5) %>%
  filter(grepl("^HALLMARK",pathway)) %>%
  mutate(pathway=sub("HALLMARK_","",pathway)) %>%
  mutate(pathway=tolower(gsub("_"," ",pathway)))
  
wb1 <- openxlsx::buildWorkbook(df1)
openxlsx::saveWorkbook(wb1,file.path(fig_dir,"Table_S2_1_pathways_gene_modules_v3.xlsx"),overwrite=T)

```
