---
title: "APM_CyCIF_prep"
author: "Kenichi Shimada"
date: '2023-02-16'
# output: 
#   html_notebook: 
#     toc: yes
---

# ERS-APM/TC CyCIF analysis

## Initial setup ---------------------------

### Install CycifAnalyzeR after updating the package (required only when)
```{r reinstall_CyCIF_ library,eval=F}
## this following code runs only when CycifAnalyzeR is not installed
if(!"CycifAnalyzeR" %in% installed.packages()){
   ## rgeos (outdated package) should be installed from source
  install.packages("https://cran.r-project.org/src/contrib/Archive/rgeos/rgeos_0.6-4.tar.gz",
                   repos=NULL,type="source")
  devtools::install_github("kenichi-shimada/CycifAnalyzeR@v1.0",force=TRUE)
}
```


### Load packages

```{r load_packages,message=F,warning=F}
# Suppress package startup messages for cleaner output
suppressPackageStartupMessages({
  library(MASS)         # For statistical functions
  library(gplots)       # For various plotting tools
  library(RColorBrewer) # For color palettes
  library(mclust)       # For Gaussian mixture models and clustering
  library(openxlsx)     # For Excel file operations
  library(dplyr)        # For data manipulation
  library(tibble)
  library(ggplot2)      # For data visualization
  library(ggrepel)      # For non-overlapping text labels in ggplot2
  library(heatmap3)     # For heatmap generation
})
```

### Set up paths
```{r}
obj_dir1 <- file.path(obj_dir,"DFBCC_cycif")
fig_dir1 <- file.path(fig_dir,"17_DFBCC_cycif_prep")
```

### Load quantification csv files into a `CycifStack` object
```{r dir_path}
if(0){
  fns1 <- dir(path=quant_data_dir,pattern=".+cellRing\\.csv")[1:3]
  cs1 <- CycifStack(fns1,path=quant_data_dir,mask_type="cellRing",mcmicro=FALSE)
  saveRDS(cs1,file.path(obj_dir1,"cs_073023.rds")) 
}else{
  cs1 <- readRDS(file.path(obj_dir1,"cs_073023.rds"))  
}

```

## set nCycles ---------------------------

### Use up to 6th cycles
```{r}
# Define the number of cycles applicable to all samples and filter samples accordingly
n.cycle <- 6
smpls <- names(cs1)
is.used.smpls <- nCycles(cs1) >= n.cycle # 29 samples

cs1 <- cs1[is.used.smpls]
nCycles(cs1) <- n.cycle
```

## Import ROIs from Omero ---------------------------
### Prepare ROIs exported from OMERO
```{r}
# Define directory for ROIs and load them
rois_dir <- file.path(cycif_data_dir,"rois")
fns <- dir(rois_dir, pattern="csv$")
names(fns) <- sub("-.+$", "", fns)

# Process and import ROIs, conditional logic included for demonstration
# Specific implementation details omitted for brevity

nm.fns <- names(fns)

rois.out <- lapply(fns,function(fn){
  tmp <- suppressWarnings(read.csv(file.path(rois_dir,fn)))
  tmp <- tmp[tmp$type != "Ellipse",]
})

# Make sure rois.out and cs1 share the identical samples
used.smpls <- intersect(names(rois.out),names(cs1))  
rois.out <- rois.out[used.smpls]
cs1 <- cs1[used.smpls]

if(0){
  ## << get df.gs2 from the bottom >>
  ids <- df.gs2$id
  names(ids) <- df.gs2$lspid
  
  is.used <- names(rois.out) %in% df.gs2$lspid
  rois.out <- rois.out[is.used]
  names(rois.out) <- as.character(ids[names(rois.out)])
}  

# batch import of ROIs to cs object
cs1 <- importROIs(cs1,rois.out)

if(0){
  # per sample ROI import
  for(smpl in c("15643","15658")){
    cs1[[smpl]] <- importROIs(cs1[[smpl]],rois.out[[smpl]])
  }
}
```

### Fig : show imported ROIs in each image
```{r}
# Loop through samples and generate ROI images
# This block contains example code for generating images; actual plotting code is not executed

smpls <- names(cs1)

for(smpl in smpls[-1]){
  tmp <- cs1[[smpl]]
  wr <- factor(as.numeric(tmp@within_rois)+1,labels=c("no","yes"))
  uniq.cols <- c("grey80","red")
  names(uniq.cols) <- levels(wr)
  png(file.path(fig_dir1,"prep_rois",paste0("rois_",smpl,".png")),width=700,height=700)
  slidePlot(tmp,
            plot_type="custom",
            custom_labs=wr,
            uniq.cols=uniq.cols,
            ttl=paste0(smpl," (up to cycle 6)"),
            ncells=1e4)
  dev.off()
}
```

## `dnaFilter` - remove lost/folded area ---------------------------

Set a filter for each channel using each cycle's DNA staining. This has to be done outside R notebook (ie regular R file) and po    sitive ROI selection.

```{r dnaFiltered_data,eval=F}
smpls <- names(cs1)

if(0){
  cs2 <- lapply(smpls,function(smpl){
     readRDS(file.path(obj_dir1,"dna_filtered",paste0(smpl,".rds")))
  })
  
  for(smpl in smpls){
    cs1[[smpl]]@dna_thres <- cs2[[smpl]]@dna_thres
  }
}

for(smpl in smpls){
  cs1[[smpl]] <- dnaFilter(cs1[[smpl]],show.only=T) # if dna.thres is available, used_cells are computed.
  # saveRDS(cs1[[smpl]],file.path(obj_dir1,"dna_filtered",paste0(smpl,".rds")))
}

```

### Fig: checking dnaFilter - # available cells on slides

```{r available_cells_slide}
# for(smpl in smpls){
#   fn <- paste0("used_cells_",smpl,"_v1_082223.png")
#   png(file.path(fig_dir1,"pre_available_cells",fn),width=1500,height=1500)
#   ttl1 <- paste0(smpl," (available cells)")
#   plotAvailCellOnSlide(cs1[[smpl]],main=ttl1,mfrow=c(4,3),
#                        legend.cex=4,cex.main=4,cex.title=4)
#   dev.off()
# }
```

### Fig : checking dnaFilter - # cells after each cycle 

```{r fig_dnaFilter}
for(smpl in smpls){
  fn <- paste0("dnaFilter_",smpl,"_v2.pdf")
  pdf(file.path(fig_dir1,"prep_dna-filter",fn),width=7,height=7)
  dnaFilter(cs1[[smpl]],show.only=T)
  dev.off()
}  
```

### Subsetting samples based on `nCycles` and `% available cells`

```{r sub_setting_cells}
## subsetting samples
stats <- cyApply(cs1,statUsedCells)
stats1 <- do.call(rbind,lapply(stats,function(st){
  st <- round(st,2)
  return(st)  
}))

colnames(stats1) <- paste0("Cycle",1:6)

stats1 <- as.data.frame(stats1) %>% tibble::rownames_to_column("id")
stats1
```

### Rename antibodies

```{r}
abs <- cs1@abs_list$ab
abs <- sub("_.+","",abs)
abs[abs=="GRANZYMEB"]<- "GrB"
abs[abs=="HLADBP1"]<- "HLADPB1"
cs1@abs_list$ab <- abs

##
for(smpl in smpls){
  cs1[[smpl]]@abs_list$ab <- abs
  names(cs1[[smpl]]@raw) <- abs
}

if(0){
  saveRDS(abs,file.path(obj_dir1,"gates","abs_101223.rds"))
}
```

## Cell-type calling ---------------------------

### Load cell-type definitions
```{r load_cell_type_def}
# ctype.file <- file.path(cycif_cycif_proj_dir,"cell-type-def","cell-type-definition_EribPem_v1.xlsx")
ctype.file <- file.path(cycif_data_dir,"cell-type-state-definition_APMTC_v2.xlsx")
ctype.sheets <- getSheetNames(ctype.file)
ctype <- readWorkbook(ctype.file,sheet="cell lineage",colNames=TRUE,rowNames=FALSE)
ctype[is.na(ctype)] <- ""
cstate <- readWorkbook(ctype.file,sheet="cell state",colNames=TRUE,rowNames=TRUE)
```

### cell type hierarchy
```{r}
# this should be included in CellTypeSkeleton
ctype.1 <- ctype[,names(ctype) %in% c("Parent","Child",abs_list(cs1)$ab)]
# used.cts <- apply(as.matrix(ctype.1[,-c(1:2)]),1,function(x)any(x %in% c("AND","OR"))) |
#                     grepl("unknown",ctype.1$Child) |
#                     grepl("other",ctype.1$Child)
# ctype.1 <- ctype.1[used.cts,]
cstate.1 <-cstate[ctype.1$Child,names(cstate) %in% abs_list(cs1)$ab]

##
pdf(file.path(fig_dir1,"prep_abs_summary","ab_graph_cyc6.pdf"),width=7,height=7)
par(mar=c(0,0,3,0))
ct.graph <- CellTypeGraph(ctype.1,plot=T,main='Cell type definition',with.hierarchy=TRUE)
dev.off()

```

### Compile and load gates, `logTh_transformation`
```{r}
if(0){
  source(file.path(cycif_src_dir,"compile_gates_101223.r"))
}else{
  gates <- readRDS(file.path(obj_dir1,"gates","gates_101223.rds")) # 120122,122822,123122  
}

if(0){
  ## to redo setGates
  for(smpl in names(cs1)){
    cs1[[smpl]]@abs_list <- cs1[[smpl]]@abs_list[1:2]
  }
}

used.smpls <- intersect(names(cs1),sub("gates_","",names(gates)))
cs1 <- setGates(cs1,gates,run_normalize=TRUE) # avoid overwrite

getGates(cs1)
```

### Cell type composition - all cells

```{r define cell types cs7} 
cs1 <- defineCellTypes(cs1,ctype.1,cstate.1,ct_name="default",overwrite=TRUE)
if(0){
  for(smpl in c("15643","15658")){
    cs1[[smpl]] <-  defineCellTypes(cs1[[smpl]],ctype.1,cstate.1,ct_name="default",overwrite=TRUE)
  }
}
nms <- names(cs1)

```

## Normalizaiton: Log-transformation -----------------
```{r}
smpls <- names(cs1)
for(smpl in names(cs1)){
  cs1[[smpl]] <- normalize(cs1[[smpl]],method="log")
}

if(0){ # save before changing the sample names
  saveRDS(cs1,file=file.path(obj_dir1,"cs1_cts_original_ids_101223.rds"))  # 27 samples
  saveRDS(cs1,file=file.path(obj_dir1,"cs1_cts_without_ellipsoid_120723.rds"))  # 27 samples
}else{
  cs1 <- readRDS(file=file.path(obj_dir1,"cs1_cts_original_ids_101223.rds"))  # 27 samples
}
```
## slidePlot : checking the cell types on slides ---------------------------

### define cell-type cols 

```{r}

```
### all cell types

```{r}
for(smpl in smpls){
  fn <- paste0(smpl,"_v2.png")
  png(file.path(fig_dir1,"cts_expression","slides","celltypes_all",fn),width=700,height=700)
  slidePlot(cs1[[smpl]],plot_type="cell_type",ncells=1e5,uniq.cols=ct.cols)
  dev.off()
}
```

### individual cell types
```{r}
for(smpl in smpls[-1]){
  d <- file.path(fig_dir1,"cts_expression","slides","celltypes_individual",smpl)
  dir.create(d)
  for(ct in uniq.cts[1:14]){
    uc <- "red"
    names(uc) <- ct
    
    fn <- paste0(smpl,"_",ct,".png")
    png(file.path(d,fn),width=1500,height=1500)
    slidePlot(cs1[[smpl]],plot_type="cell_type",ncells=1e5,uniq.cols=uc)
    dev.off()
  }
}
```

## Metadata - change Match sample names -----------------
```{r}
## BOT <-> LSP
hrtme <- read.xlsx(file.path(cycif_data_dir,"CyCIF_HRTME_ANALYSISTRACKING.xlsx"),1)

lsp2bot <- hrtme[1:2] %>%
  rename(lspid = `LSP.ID.#`) %>%
  rename(bot = `BOT.#`) %>%
  mutate(lspid = as.character(lspid)) %>%  
  mutate(bot = as.character(bot)) %>%
  arrange(lspid)
  

bot.cycif <- hrtme$`BOT.#`
names(bot.cycif) <- hrtme$`LSP.ID.#`

table(names(cs1) %in% names(bot.cycif)) ## all TRUE

## BOT <-> TXN: ERS-APMTC scores
df.gs <- readRDS(file.path(obj_dir,"DFBCC_txn","dfbcc_bulk_mod_activity.rds"))
df.gs1 <- df.gs %>%
  arrange(desc(ERS)) %>%
  dplyr::rename(id="new") %>%
  mutate(bot=sub("BOT","",bot)) %>%
  left_join(lsp2bot,by="bot") %>%
  filter(lspid != "15631") %>%
  arrange(lspid)

df.gs1 <- df.gs1[c("id",names(df.gs1[-6]))]

for(i in seq(nrow(df.gs1))){
  cs1[[i]]@cell_types$default@sample_names[] <- df.gs1$id[i]
}
names(cs1) <- df.gs1$id
pData(cs1) <- df.gs1

o <- order(as.numeric(sub("df","",df.gs1$id)))
cs1 <- cs1[o]

names(names(cs1)) <- c()
```

```{r}
## modify cell_state_marker table
nms <- names(cs1)

for(nm in nms){
  x <- cs1[[nm]]
  csts <- x@cell_types$default@cell_state_def[,c(1:2,4)] # 1:7 is the cell state markers (PD1 is duplicated)
  csts <- csts %>%
    mutate(PD1=c(NA,"CAN")[(rownames(csts) %in% c("T","CD8T","CD4T","CD4T_nonTreg","CD4T_Treg","T_other")) + 1]) %>%
    mutate(across(everything(), ~dplyr::na_if(.x, ""))) %>%
    tibble::rownames_to_column(var = "cell_types")
  x@cell_types$default@cell_state_def <- csts 
  cs1[[nm]] <- x
}
```

```{r}
df.gs1 <- pData(cs1) %>% arrange(as.numeric(sub("df","",id)))
pData(cs1) <- df.gs1
```

## Module expression `df.gs`
```{r}
# cor(m,m) # column-wise comparison
df.gs.1 <- pData(cs1) %>% select(1,3:6)
rownames(df.gs.1) <- names(cs1)

df.gs <- df.gs.1[-1]
uniq.mods <- names(df.gs)
```

## Save `cs1` - after identifying `cell types`
```{r define cell types} 
if(0){
  saveRDS(cs1,file=file.path(obj_dir1,"cs.rds"))  # 28 samples
}else{
  cs1 <- readRDS(file=file.path(obj_dir1,"cs.rds"))
}
```


## Cell type frequency

### compute cell type frequency
```{r}
nsh <- cellTypeFrequency(cs1,simple=TRUE)
nsh1 <- t(nsh) %>% as.data.frame() %>% rownames_to_column("id")
uniq.cts <- names(nsh)[-1]
```

### [Fig 4C] cell type frequency
```{r}
pdf(file.path(fig_dir1,"cts","all_celltypes_sorted.pdf"),width=9,height=7)
par(mar=c(5,5,4,10))
barplot(nsh[14:1,],col=rev(ct.cols),border=FALSE,las=2,main="Cell type frequency")
legend(par()$usr[2],par()$usr[4],legend=names(ct.cols),fill=ct.cols,xpd=TRUE,border = "white",cex=.8)
dev.off()
```

### [Fig 4D] correlation between cell type frequency
```{r}
cor.cts <- cor(t(nsh))
pdf(file.path(fig_dir1,"cts","cts_heatmap.pdf"),width=7,height=7)
heatmap3(cor.cts,balanceCol=TRUE,scale="none",margins=c(10,10),
         main="Cell type frequency correlation (CyCIF - bulk)")
dev.off()
```

### [Fig 4E] correlation between cell type frequency and module activity
```{r}
cor.mat <- cor(t(nsh),df.gs)

## barplot for each module
rng <- c(-1,1)*.85 #(cor.mat)
for(i in 1:4){
  this.cor <- sort(cor.mat[,i])
  pdf(file.path(fig_dir1,"cts",paste0("module",i,".pdf")),width=4,height=7)
  par(mar=c(5,10,4,1))
  barplot(this.cor,col=ct.cols[names(this.cor)],horiz=TRUE,main=names(df.gs)[i],las=1,xlim=rng)
  abline(v=c(.4,-.4),lty=2)
  box()
  dev.off()
}
```

```{r}
saveRDS(nsh,file=file.path(obj_dir1,"nstat_cycif.rds"))
```


