---
title: "11_ACOSOG_prep.Rmd"
# output: html_notebook
---

## Initial setting
### Load libraries
```{r}
suppressPackageStartupMessages({
  library(GEOquery)
  # library(limma)
  library(ggrepel)
  library(RColorBrewer)
  library(dplyr)
  library(GSVA)
  library(impute)
  library(tidyr)
  library(heatmap3)
})
```

### Set path
```{r}
fig_dir1 <- file.path(fig_dir,"22_ACOSOG_Z1031B")
```

### Load data
#### txn data - from GEO and gene modules
```{r}
Sys.setenv("VROOM_CONNECTION_SIZE" = 131072 * 5)
gset <- getGEO("GSE87411", GSEMatrix = TRUE, AnnotGPL=FALSE)
if (length(gset) > 1) idx <- grep("GPL6480", attr(gset, "names")) else idx <- 1

gset <- gset[[idx]]
```

#### module genes (`mods`)
```{r}
mods <- readRDS(file=file.path(obj_dir,"Gene_modules","module_genes_final.rds"))
```

### DE analysis functions
```{r}
# Create function for gene categorization
DE_category <- function(x,gs){
  x$DE_genes <- "others"
  for(i in seq(gs)){
    nm <- names(gs)[i]
    gs1 <- gs[[i]]
    # x$DE_genes[x$adj.P.Val < 0.05 &  x$GENE_SYMBOL %in% gs1] <- nm
    x$DE_genes[x$GENE_SYMBOL %in% gs1] <- nm    
  }
  x$DE_genes <- factor(x$DE_genes,levels=c(names(gs),"others"))
  return(x)
}

#Create plotting function
volcano_plot<- function(df,p.thres=1e-5){
  ## set my colors
  # bl <- colorRampPalette(c(brewer.pal(3,"Dark2")[3],"white"))(4)
  # bl <- colorRampPalette(c("white","steelblue"))(6)[5]
  bl <- "steelblue"
  mycols <- c(bl,brewer.pal(3,"Dark2"),"grey80")
  names(mycols) <- c(names(mods),"others")

  df <- df %>%
    mutate(text=ifelse((adj.P.value < p.thres & Gene_category != "others") | 
                         (GENE_SYMBOL %in% c("ESR1","CD8A","HLA-A")),GENE_SYMBOL,""))
  p <- ggplot(df %>% arrange(desc(Gene_category)),
              aes(x=logFC,y=-log10(adj.P.value),label=text)) +
    geom_point(aes(color=Gene_category,size=Gene_category)) +
    geom_hline(yintercept=-log10(0.05),color="grey60") +
    scale_color_manual(values=mycols) +
    scale_size_manual(values=c(1.5,1.5,1.5,1.5,.5)) +
    geom_text_repel(max.overlaps=Inf,size=2) +
    theme_minimal() 
  return(p)
}
```

## Update metadata
### Add TimePoints
```{r}
# Time points to measure transcriptome
pd.need <- pData(gset) %>% 
  mutate(time_point = factor(as.character(rep(0:1,times=109)),labels=c("Pre","Post"))) %>%
  dplyr::select(`studynum:ch2`,title,time_point,`bl  ki67:ch2`,`eot ki67:ch2`,`endocrine therapy response group:ch2`,`treatment course:ch2`) %>%
  tibble::rownames_to_column("id")  
names(pd.need) <- c("id","patient","title","time_point","bl_ki67","eot_ki67","response_group","treatment_course")

pd.need1 <- pd.need %>%
  mutate(time_point=factor(time_point,levels=c("Pre","Post"))) %>%
  mutate(response_group=factor(response_group,levels=c("resistant","sensitive"))) %>%
  mutate(bl_ki67=ifelse(bl_ki67=="60-70%","65%",bl_ki67)) %>%
  mutate(bl_ki67=as.numeric(sub("~?([\\.0-9]+)%?","\\1",bl_ki67))) %>%
  mutate(eot_ki67=ifelse(eot_ki67=="<1%","0.5%",eot_ki67)) %>%
  mutate(eot_ki67=ifelse(eot_ki67==">90%","95%",eot_ki67)) %>%  
  mutate(eot_ki67=as.numeric(sub("[~]?([\\.0-9]+)%?","\\1",eot_ki67))) %>%
  filter(!is.na(bl_ki67) & !is.na(eot_ki67)) %>%
  mutate(patient = factor(patient))

```

### Refine `response_group`
```{r}
pd.need1 <- pd.need1 %>% 
  mutate(ratio = eot_ki67/bl_ki67) %>%
  mutate(diff = eot_ki67 - bl_ki67) %>%
  filter(!is.na(ratio) & !is.na(diff))

# n <- sum(pd.need1$diff > 0)
n <- sum(pd.need1$ratio > 1)
pos.th <- sort(pd.need1$ratio,decreasing=T)[n]
neg.th <- sort(pd.need1$ratio,decreasing=F)[n]

pos <- (pd.need1 %>% filter(ratio >= pos.th))$id
neg <- (pd.need1 %>% filter(ratio <= neg.th))$id

labs <- c("sensitive","intermediate","resistant")[(pd.need1$ratio >= pos.th) + (pd.need1$ratio > neg.th) + 1]

pd.need1 <- pd.need1 %>%
  mutate(my_response1 = factor(labs,levels=c("sensitive","intermediate","resistant")))
rownames(pd.need1) <- pd.need1$id

pData(gset) <- pd.need1
```

#### [not-used] plot - Ki-67 status pre and post
```{r}
ptcols <- c(brewer.pal(3,"Set1"),"grey80")[c(1,4,2)]
names(ptcols) <- levels(pd.need$my_response1)

p <- pd.need1 %>% 
  arrange(desc(my_response1)) %>%
  ggplot(aes(x=bl_ki67,y=eot_ki67,color=my_response1,size=my_response1)) +
  theme_minimal() +
  geom_vline(xintercept=0) +  
  geom_hline(yintercept=0) +
  geom_abline(slope=1,intercept=0) +  
  # geom_abline(slope=1,intercept=pos.th,linetype="dashed") +
  # geom_abline(slope=1,intercept=neg.th,linetype="dashed") +
  geom_abline(slope=neg.th,intercept=0,linetype="dashed") +  
  
  geom_point() +
  scale_color_manual(values=ptcols) +
  scale_size_manual(values=c(1.5,0.5,1.5)) + 
  ggtitle("Ki67 positive tumors (%) at Pre- and Post-tx") +
  xlab("Baseline") +
  ylab("End of treatment")

pdf(file=file.path(fig_dir1,"ACSOG_scatter_baseline.pdf"),width=7,height=7)
print(p)
dev.off()
```

## ssgsea analysis 
```{r}
ex <- exprs(gset)
sn <- pd.need1$id
ex <- ex[,sn]
# identical(rownames(ex),fData(gset)$ID) # TRUE
```

### impute missing values
```{r}
syms <- fData(gset)$GENE_SYMBOL
rownames(ex) <- syms

not.na <- apply(ex,1,function(x)sum(is.na(x))/length(x) < .1)
exist.syms <- syms[not.na]

mods.0 <- lapply(mods,function(x)x[x %in% exist.syms])
sapply(mods.0,length)
mods.1 <- lapply(mods,function(x)x[!x %in% exist.syms]) 
sapply(mods.0,length)

ex1 <- ex[not.na,]

sum(is.na(ex))/prod(dim(ex)) # 6% missing...
ex1 <- impute.knn(ex1,k=10)
ex1 <- ex1$data
```

### PCA & change of Ki67
```{r}
pca <- prcomp(t(ex1))

## difference in Ki67 
ki67 <- rep(NA,nrow(pd.need1))
ki67[pd.need1$time_point == "Pre"] <- pd.need1$bl_ki67[pd.need1$time_point == "Pre"]
ki67[pd.need1$time_point == "Post"] <- pd.need1$eot_ki67[pd.need1$time_point == "Post"]

d.ki67 <- pd.need1 %>% filter(time_point =="Post") %>% pull(eot_ki67) - 
  pd.need1 %>% filter(time_point =="Post") %>% pull(bl_ki67)

pd.need1 <- pd.need1[1:11] %>%
  cbind(pca$x[,1:2]) %>%
  mutate(ki67=ki67) %>%
  mutate(d.ki67 = rep(d.ki67,each=2))

if(0){
  cols <- (as.numeric(pd.need1$my_response1) + c(as.numeric(pd.need1$time_point)-1)*3)
  plot(pca$x[,1],pca$x[,2],col=brewer.pal(6,"Paired")[cols],pch=19)

  ## the trend was not too strong.
  ggplot(pd.need1,aes(x=PC1,y=PC2,color=my_response1,shape=time_point)) + 
    geom_point() +
    geom_line(aes(group=patient)) +
    theme_minimal()
  
  ggplot(pd.need1,aes(x=PC1,y=PC2,color=d.ki67,shape=time_point)) + 
    geom_point() +
    geom_line(aes(group=patient)) +
    scale_color_gradient2(low="blue",mid="black",high="red",midpoint=0) +
    theme_minimal()
}
```

### ssGSEA, 4 modules
```{r}
syms1 <- rownames(ex1)
mods1 <- lapply(mods,function(x)x[x %in% syms1])
sapply(mods1,length)

system.time({
  ssgsea.par <- ssgseaParam(expr = ex1, geneSets = mods1, minSize = 10, maxSize = 500, normalize=FALSE)
  scores <- gsva(ssgsea.par)
  scores <- t(scores)
})

scores <- scores %>% as.data.frame() %>% tibble::rownames_to_column("id")
pd.need1 <- pd.need1[1:15] %>% left_join(scores,by="id")

d.exp <- pd.need1 %>% 
  dplyr::select(patient,time_point,!!!syms(names(scores)[-1]))

d.exp1 <- (d.exp %>% filter(time_point =="Post") %>% dplyr::select(-(1:2)) -
  d.exp %>% filter(time_point =="Pre") %>% dplyr::select(-(1:2))) %>%
  cbind(d.exp %>% filter(time_point =="Pre") %>% dplyr::select(1))
names(d.exp1)[1:4] <- paste0("d.",names(d.exp1)[1:4])
# d.exp1$d.ERS[d.exp1$d.ERS < -3000] <- -3000
# d.exp1$d.ERS[d.exp1$d.ERS > 3000] <- 3000

pd.need2 <- pd.need1 %>% left_join(d.exp1,by="patient")
```

### [Fig. 7A] comparison between ERS_Pre and d.ERS
```{r}
ers <- pd.need2 %>% 
  select(patient,time_point,`ERS`) %>%
  spread(time_point,`ERS`) %>%
  mutate(diff=Post-Pre)
pe <- round(cor(ers$Pre,ers$diff,method="pe"),2)
sp <- round(cor(ers$Pre,ers$diff,method="sp"),2)

p <- ggplot(ers,aes(x=Pre,y=diff)) +
  geom_point(color="grey20") +
  ggtitle(paste0("ERS (PCC:",pe,",SCC:",sp,")")) +
  geom_hline(yintercept=0) +
  xlab("ERS expression at Pre-treatment") +
  ylab("Changes in ERS expression during treatment") +
  theme_minimal()

pdf(file=file.path(fig_dir1,"ERS_pre_post.pdf"),width=4,height=4)
print(p)
dev.off()

```


```{r}
if(0){
  p <- pd.need2 %>% 
    select(patient,time_point,`TNFa/NFkB`) %>%
    spread(time_point,`TNFa/NFkB`) %>%
    mutate(diff=Post-Pre) %>%
    ggplot(aes(x=Pre,y=diff)) +
    geom_point() +
    ggtitle(paste0("TNFa/NFkB (SCC:",sp,")")) +
    xlab("TNFa/NFkB exprs. at Pre-tx") +
    ylab("∆TNFa/NFkB") +
    theme_minimal()
  
  p <- pd.need2 %>% 
    select(patient,time_point,`IFN-I`) %>%
    spread(time_point,`IFN-I`) %>%
    mutate(diff=Post-Pre) %>%
    ggplot(aes(x=Pre,y=diff)) +
    geom_point() +
    ggtitle(paste0("IFN-I (SCC:",sp,")")) +
    xlab("IFN-I exprs. at Pre-tx") +
    ylab("∆IFN-I") +
    theme_minimal()
  
  p <- pd.need2 %>% 
    select(patient,time_point,`APM/TC`) %>%
    spread(time_point,`APM/TC`) %>%
    mutate(diff=Post-Pre) %>%
    select(Pre,diff) %>%
    cor(.,method="spearman")
    ggplot(aes(x=Pre,y=Post)) +
    geom_point() +
    ggtitle(paste0("APM/TC (SCC:",sp,")")) +
    xlab("APM/TC exprs. at Pre-tx") +
    ylab("∆APM/TC") +
    theme_minimal()
}
```

## Module expression changes
### [Fig 7A] Violin plots
```{r}
pval <- t.test(pd.need2 %>% filter(time_point == "Pre") %>% pull(ERS),
       pd.need2 %>% filter(time_point == "Post") %>% pull(ERS),
       paired=T)$p.value
tmp <- data.frame(x=1.5,y=max(pd.need2$ERS),label=paste0("p=",signif(pval,2)))
p <- ggplot(pd.need2,aes(x=time_point,y=ERS)) +
  geom_violin() +#fill="grey80") +
  geom_text(data=tmp,aes(x=x,y=y,label=label)) +
  geom_line(aes(group=patient,color=d.ERS),linewidth=.1) +
  scale_color_gradient2(low="blue",mid="black",high="red",midpoint=0) +
  ylab("module scores") +
  ggtitle("ERS") +
  theme_minimal()
pdf(file=file.path(fig_dir1,"violin_ERS_pre_post.pdf"),width=4,height=7)
print(p)
dev.off()

  ## TNFa/NFkB
pval <- t.test(pd.need2 %>% filter(time_point == "Pre") %>% pull(`TNFa/NFkB`),
       pd.need2 %>% filter(time_point == "Post") %>% pull(`TNFa/NFkB`),
       paired=T)$p.value
tmp <- data.frame(x=1.5,y=max(pd.need2$`TNFa/NFkB`),label=paste0("p=",signif(pval,2)))
p <- ggplot(pd.need2,aes(x=time_point,y=`TNFa/NFkB`)) +
  geom_violin() +#fill="grey80") +
  geom_text(data=tmp,aes(x=x,y=y,label=label)) +
  geom_line(aes(group=patient,color=`d.TNFa/NFkB`),linewidth=.1) +
  scale_color_gradient2(low="blue",mid="black",high="red",midpoint=0) +
  ylab("module scores") +
  ggtitle("TNFa/NFkB") +
  theme_minimal()
pdf(file=file.path(fig_dir1,"violin_NFkB_pre_post.pdf"),width=4,height=7)
print(p)
dev.off()

## APM/TC
pval <- t.test(pd.need2 %>% filter(time_point == "Pre") %>% pull(`APM/TC`),
       pd.need2 %>% filter(time_point == "Post") %>% pull(`APM/TC`),
       paired=T)$p.value
tmp <- data.frame(x=1.5,y=max(pd.need2$`APM/TC`),label=paste0("p=",signif(pval,2)))
p <- ggplot(pd.need2,aes(x=time_point,y=`APM/TC`)) +
  geom_violin() +#fill="grey80") +
  geom_text(data=tmp,aes(x=x,y=y,label=label)) +
  geom_line(aes(group=patient,color=`d.APM/TC`),linewidth=.1) +
  scale_color_gradient2(low="blue",mid="black",high="red",midpoint=0) +
  ylab("module scores") +
  ggtitle("APM/TC") +
  theme_minimal()
pdf(file=file.path(fig_dir1,"violin_APMTC_pre_post.pdf"),width=4,height=7)
print(p)
dev.off()

## IFN-I
pval <- t.test(pd.need2 %>% filter(time_point == "Pre") %>% pull(`IFN-I`),
       pd.need2 %>% filter(time_point == "Post") %>% pull(`IFN-I`),
       paired=T)$p.value
tmp <- data.frame(x=1.5,y=max(pd.need2$`IFN-I`),label=paste0("p=",signif(pval,2)))

p <- ggplot(pd.need2,aes(x=time_point,y=`IFN-I`)) +
  geom_violin() +#fill="grey80") +
  geom_text(data=tmp,aes(x=x,y=y,label=label)) +
  geom_line(aes(group=patient,color=`d.IFN-I`),linewidth=.1) +
  scale_color_gradient2(low="blue",mid="black",high="red",midpoint=0) +
  ylab("module scores",size=5) +
  ggtitle("IFN-I") +
  theme_minimal()
pdf(file=file.path(fig_dir1,"violin_IFN_pre_post.pdf"),width=4,height=7)
print(p)
dev.off()
```

### [Fig. 7A] d.ERS vs d.Ki67
```{r}
pcc <- cor(pd.need2$`d.ERS`,pd.need2$`d.ki67`,method="pearson")
scc <- cor(pd.need2$`d.ERS`,pd.need2$`d.ki67`,method="spearman")

p <- ggplot(pd.need2,aes(x=d.ERS,y=d.ki67)) +
    geom_hline(yintercept=0) +
    geom_vline(xintercept=0) +
    geom_point(color="grey20") +
    ggtitle(paste0("Changes in ERS and Ki67\n(pcc=",signif(pcc,2),",scc=",signif(scc,2),")")) +
    theme_minimal()

pdf(file=file.path(fig_dir1,"ERS_ki67.pdf"),width=4,height=4)
print(p)
dev.off()

if(0){
  p <- ggplot(pd.need2,aes(x=d.ERS,y=`d.APM/TC`)) +#,color=`ki67`,shape=time_point)) + 
    geom_point() +
    geom_hline(yintercept=0) +
    geom_vline(xintercept=0) +
    geom_smooth(method="lm",se=T) +
    theme_minimal()
  
  ggplot(pd.need2,aes(x=d.ERS,y=`d.TNFa/NFkB`)) +#,color=`ki67`,shape=time_point)) + 
    geom_point() +
    geom_hline(yintercept=0) +
    geom_vline(xintercept=0) +
    geom_smooth(method="lm",se=T) +
    theme_minimal()
  
  ggplot(pd.need2,aes(x=d.ERS,y=`d.IFN-I`)) +#,color=`ki67`,shape=time_point)) + 
    geom_point() +
    geom_hline(yintercept=0) +
    geom_vline(xintercept=0) +
    geom_smooth(method="lm",se=T) +
    theme_minimal()
}
```

```{r}
if(0){
  d.pd <- pd.need2 %>% 
    filter(time_point == "Pre") %>%
    dplyr::select(patient,time_point,14:23)
  # ERS,`TNFa/NFkB`,`APM/TC`,`IFN-I`,
  #                 d.ERS,`d.TNFa/NFkB`,`d.APM/TC`,`d.IFN-I`,d.ki67,bl_ki67,eot_ki67)
  d.pd$d.ERS[d.pd$d.ERS < -3000] <- -3000
  d.pd$d.ERS[d.pd$d.ERS > 3000] <- 3000
  
  heatmap3(cor(d.pd %>% dplyr::select(-patient,-time_point),method="sp"),
           scale="none",balanceColor=TRUE,margins=c(10,10))
  
  lm1 <- lm(`d.APM/TC` ~ `d.IFN-I` + `d.ERS` + `d.TNFa/NFkB`,data=d.pd)
  
  ggplot(d.pd,aes(x=`IFN-I`,y=`d.APM/TC`,color=d.ERS)) +
    geom_point(alpha=0.5) + 
    scale_color_gradient2(low="blue",mid="black",high="red",midpoint=0) +
    geom_line(aes(group=patient)) +
    theme_minimal()
  
  ggplot(d.pd,aes(x=bl_ki67,y=eot_ki67,color=d.ERS)) +
    geom_point(alpha=0.5) + 
    scale_color_gradient2(low="blue",mid="black",high="red",midpoint=0) +
    geom_line(aes(group=patient)) +
    theme_minimal()
  
  ggplot(d.pd,aes(x=bl_ki67,y=eot_ki67,color=d.ERS)) +
    geom_point(alpha=0.5) + 
    scale_color_gradient2(low="blue",mid="black",high="red",midpoint=0) +
    geom_line(aes(group=patient)) +
    theme_minimal()
}
```


### [Fig 7BC] correlation
```{r}
pre <- cor(pd.need2 %>% filter(time_point == "Pre") %>% dplyr::select(16:19),method="sp")
pdf(file=file.path(fig_dir1,"scc_pre.pdf"),width=7,height=7)
heatmap3(pre,scale="none",balanceColor=T,main="Pre",margins=c(10,10))
dev.off()
write.csv(round(pre,2),file=file.path(fig_dir1,"scc_pre.csv"))

##
changes <- cor(pd.need2 %>% filter(time_point == "Post") %>% dplyr::select(20:23),method="sp")
pdf(file=file.path(fig_dir1,"scc_change.pdf"),width=7,height=7)
heatmap3(changes,scale="none",balanceColor=T,main="change",margins=c(10,10))
dev.off()
write.csv(round(changes,2),file=file.path(fig_dir1,"scc_change.csv"))

##
pre.post <- cor(pd.need2 %>% filter(time_point == "Pre") %>% dplyr::select(16:19),
                    pd.need2 %>% filter(time_point == "Pre") %>% dplyr::select(c(20:23)),method="sp")
heat.cols <- colorRampPalette(c("navy", "white", "firebrick3"))(201)
names(heat.cols) <- -100:100
n <- round(max(abs(range(pre.post)))*100)

heat.cols <- heat.cols[as.character(seq(-n,n))]

## 
pdf(file=file.path(fig_dir1,"scc_pre_post.pdf"),width=7,height=7)
heatmap3(pre.post,margins=c(10,10),col=heat.cols,
    Colv=NA,Rowv=NA,
    scale="none",balanceColor=T,main="correlation between initial state and the change")
dev.off()
write.csv(round(pre.post,2),file=file.path(fig_dir1,"scc_pre_post.csv"))
# "#7FC97F" "#FDC086" "#BEAED4"
```
### [Fig 7D] linear regression
```{r}
pd.need2.lm <- pd.need2 %>% filter(time_point == "Pre") %>%
  mutate(`APM/TC` = scale(`APM/TC`)) %>%
  mutate(`IFN-I` = scale(`IFN-I`)) %>%
  mutate(`TNFa/NFkB` = scale(`TNFa/NFkB`)) %>%
  mutate(`d.IFN-I` = scale(`d.IFN-I`)) %>%
  mutate(`d.TNFa/NFkB` = scale(`d.TNFa/NFkB`)) %>%
  mutate(`d.APM/TC` = scale(`d.APM/TC`))

lm1 <- lm(`d.APM/TC` ~ `APM/TC` + `IFN-I` + `TNFa/NFkB` + `d.IFN-I` + `d.TNFa/NFkB`,
          data=pd.need2.lm)
nlps <- sort(-log10(summary(lm1)$coef[-1,4]))
names(nlps) <- gsub("`","",names(nlps))
lm.coefs <- summary(lm1)$coef[-1,1]

cols <- mod.uniq.cols[sub("^d\\.","",names(nlps))]
pdf(file=file.path(fig_dir1,"lm_nlps_dAPMTC.pdf"),width=7,height=4)
par(mar=c(5,10,4,1))
barplot(nlps,las=1,horiz=T,xlab="-log10(p-value)",main="significance of the coefficients",col=cols)
abline(v=-log10(0.05),lty=2)
dev.off()

if(0){
  pdf(file=file.path(fig_dir1,"lm_coef_dAPMTC.pdf"),width=7,height=4)
  par(mar=c(5,10,4,1))
  barplot(lm.coefs[names(nlps)],las=1,horiz=T,xlab="coefficient",main="contribution of the coefficients")
  abline(v=0)
  dev.off()
}

```

## cell type deconv

### biomarkers for different cell types
```{r}
markers.wd <- readRDS(file=file.path(obj_dir,"celltype_markers","markers_all_wehi_dfbcc.rds"))

th <- 1e-100
m1 <- markers.wd %>%
  filter(p_val_adj.d < th & p_val_adj.w < th)
uniq.gns <- names(which(table(m1$gene)<3))
m1 <- m1 %>% filter(gene %in% uniq.gns)

syms1 <- rownames(ex1)
lst.wd <- tapply(m1$gene,m1$cluster,function(x){
  # x <- x[x %in% syms1 & !x %in% unlist(mods)]
  x <- x[x %in% syms1]
  n <- min(length(x),100)
  # n <- length(x)
  return(x[seq(n)])
})
lst.wd <- lapply(lst.wd,function(x)x)
sapply(lst.wd,length)
```


```{r}
sapply(lst.wd,function(x)length(intersect(x,unlist(mods)))) # overlap between cell type markers and modules - still keep them
```

### ssgsea to compute relativea abundance
```{r}
system.time({
  # ssgsea.par <- ssgseaParam(expr = ex[,1:5], geneSets = mods, minSize = 10, maxSize = 500)
  ssgsea.par <- ssgseaParam(expr = ex1, geneSets = lst.wd, minSize = 10, maxSize = 1000, normalize=FALSE)
  scores <- gsva(ssgsea.par)
  scores <- t(scores)
})
colnames(scores)[1] <- "Tumor"
pd.need3 <- pd.need2 %>% cbind(scores)
pd.need3pre <- pd.need3 %>% filter(time_point == "Pre")

pe <- cor(pd.need3pre[16:19],pd.need3pre[24:35],method="pearson")
sp <- cor(pd.need3pre[16:19],pd.need3pre[24:35],method="spearman")

sort(pe["TNFa/NFkB",],dec=T)
sort(sp["TNFa/NFkB",],dec=T)

sort(pe["IFN-I",],dec=T)
sort(sp["IFN-I",],dec=T)

d.scores <- scores[rep(1:2,times=107)==2,] - scores[rep(1:2,times=107)==1,]
colnames(d.scores) <- paste0("d.",colnames(scores))

d.pe <- cor(pd.need3 %>% filter(time_point == "Pre") %>% dplyr::select(20:23),d.scores,method="pearson")
d.sp <- cor(pd.need3 %>% filter(time_point == "Pre") %>% dplyr::select(20:23),d.scores,method="spearman")

sort(d.pe["d.TNFa/NFkB",],dec=T)
sort(d.sp["d.TNFa/NFkB",],dec=T)

plot(pe["TNFa/NFkB",],sp["TNFa/NFkB",])
plot(d.pe["d.TNFa/NFkB",],d.sp["d.TNFa/NFkB",])

sp.te <- cor(pd.need3pre$`TNFa/NFkB`,pd.need3pre$Endo,method="spearman")
d.sp.te <- cor((pd.need3 %>% filter(time_point=="Pre"))$`d.TNFa/NFkB`,d.scores[,"d.Endo"],method="spearman")
```


```{r}
p <- ggplot(pd.need3pre,aes(`TNFa/NFkB`,Endo)) +
  geom_point() +
  xlab("TNFa/NFkB")+
  ylab("Endo")+
  ggtitle(paste0("SCC btwn TNFa/NFkB and Endo\n(SCC=",signif(sp.te,2),")")) +
  theme_minimal()

pdf(file=file.path(fig_dir1,"scc_nfkb_endo_pre.pdf"),width=4,height=4)
print(p)
dev.off()
##
pd.need3.d <- pd.need3 %>% 
  filter(time_point=="Post") %>%
  left_join(as.data.frame(d.scores) %>% tibble::rownames_to_column("id"),by="id")
  
p <- ggplot(pd.need3.d, aes(`d.TNFa/NFkB`,`d.Endo`)) +
  geom_point() +
  theme_minimal() +
  geom_hline(yintercept=0) +
  geom_vline(xintercept=0) +
  ggtitle(paste0("SCC btwn d.TNFa/NFkB and d.Endo\n(SCC=",signif(d.sp.te,2),")"))

pdf(file=file.path(fig_dir1,"scc_dnfkb_dendo_v2.pdf"),width=4,height=4)
print(p)
dev.off()

# heatmap3(pe,scale="none",balanceColor=TRUE,margins=c(10,10))
pdf(file=file.path(fig_dir1,"scc_mod_cts_pre_v2.pdf"),width=7,height=7)
heatmap3(sp,scale="none",balanceColor=TRUE,margins=c(10,10),
         main="correlation between module activity and cell types")
dev.off()

# heatmap3(d.pe,scale="none",balanceColor=TRUE,margins=c(10,10))
pdf(file=file.path(fig_dir1,"scc_mod_cts_change_v2.pdf"),width=7,height=7)
heatmap3(d.sp,scale="none",balanceColor=TRUE,margins=c(10,10),
         main="correlation between changes in module activity and cell types")
dev.off()
```

## DEG analysis 
### Comparison of the baseline between AI-resistant and sensitive groups
```{r}
ex <- exprs(gset)
syms <- fData(gset)$GENE_SYMBOL
```

```{r}
pd.need.1 <- pd.need1 %>% 
  # mutate(my_response2 = c("sensitive","resistant")[1 + (d.ki67 > 0)]) %>%
  # mutate(my_response2 = factor(my_response2))
  filter(my_response1 %in% c("sensitive","resistant")) %>%
  mutate(my_response1 = factor(my_response1))
pd.need.pre <- pd.need.1 %>% filter(time_point == "Pre")

clin.ids <- tapply(pd.need.pre$id,pd.need.pre$my_response1,identity)

out.bl <- data.frame(t(apply(ex,1,function(x1){
  lfc <- mean(x1[clin.ids$sensitive],na.rm=T) - mean(x1[clin.ids$resistant],na.rm=T)
  p.val <- try(t.test(x1[clin.ids$sensitive],x1[clin.ids$resistant],paired=FALSE)$p.value,silent=T)
  if(class(p.val)=="try-error"){
    p.val <- NA
  }
  return(c(lfc=lfc,p.val=p.val))
}))) %>%
  mutate(nlp=-log10(p.val)) %>%
  mutate(GENE_SYMBOL=syms)

fdr <- qvalue::qvalue(out.bl$p.val)$qvalues
out.bl <- out.bl %>% 
  mutate(qval = fdr) %>%
  mutate(nlq = -log10(qval)) %>%
  mutate(slq = sign(lfc) * nlq)

## remove redundant genes
idx <- tapply(seq(syms),syms,function(i){
  tmp <- out.bl$nlq[i]
  max.i <- i[which.max(tmp)]
  if(length(max.i)==0){
    max.i <- NA
  }
  return(max.i)
},simplify=T)

out.bl <- out.bl[idx,]
out.bl <- out.bl[-1,]

out.bl <- DE_category(out.bl,gs=mods)
names(out.bl) <- c("logFC","P.value","-log10(P.value)","GENE_SYMBOL","adj.P.value","-log10(adj.P.value)","signed log adj.p","Gene_category")
```

#### Plot
```{r}
pdf(file=file.path(fig_dir1,"ACSOG_volcano_baseline_all_smpls.pdf"),width=7,height=7)
volcano_plot(out.bl,p.thres=-log10(0.5)) + ggtitle("Sensitive vs Resistant (Baseline)") 
dev.off()
```

### Comparison of Pre-to-Post change between AI-resistant and sensitive groups
#### Sensitive (pre and post)
```{r}
pd.need.sen <- pd.need.1 %>% 
  filter(my_response1 == "sensitive") %>% 
  mutate(my_response1 = factor(my_response1))

clin.ids <- tapply(pd.need.sen$id,pd.need.sen$time_point,identity)

out.sen <- data.frame(t(apply(ex,1,function(x1){
  lfc <- mean(x1[clin.ids$Post],na.rm=T) - mean(x1[clin.ids$Pre],na.rm=T)
  p.val <- t.test(x1[clin.ids$Pre],x1[clin.ids$Post],paired=T)$p.value
  return(c(lfc=lfc,p.val=p.val))
}))) %>%
  mutate(nlp=-log10(p.val)) %>%
  mutate(GENE_SYMBOL=syms) 

fdr <- qvalue::qvalue(out.sen$p.val)$qvalues
out.sen <- out.sen %>% 
  mutate(qval = fdr) %>%
  mutate(nlq = -log10(qval)) %>%
  mutate(slq = sign(lfc) * nlq)

## remove redundant genes
idx <- tapply(seq(syms),syms,function(i){
  tmp <- out.sen$nlq[i]
  max.i <- i[which.max(tmp)]
  return(max.i)
},simplify=T)

out.sen <- out.sen[idx,]
out.sen <- out.sen[-1,]

out.sen <- DE_category(out.sen,gs=mods)
names(out.sen) <- c("logFC","P.value","-log10(P.value)","GENE_SYMBOL","adj.P.value","-log10(adj.P.value)","signed log adj.p","Gene_category")
```

#### Resistant (pre and post)
```{r}
pd.need.res <- pd.need.1 %>% 
  filter(my_response1 == "resistant") %>% 
  mutate(my_response1 = factor(my_response1))

clin.ids <- tapply(pd.need.res$id,pd.need.res$time_point,identity)

out.res <- data.frame(t(apply(ex,1,function(x1){
  lfc <- mean(x1[clin.ids$Post],na.rm=T) - mean(x1[clin.ids$Pre],na.rm=T)
  p.val <- t.test(x1[clin.ids$Pre],x1[clin.ids$Post],paired=T)$p.value
  return(c(lfc=lfc,p.val=p.val))
}))) %>%
  mutate(nlp=-log10(p.val)) %>%
  mutate(GENE_SYMBOL=syms) 

fdr <- qvalue::qvalue(out.res$p.val)$qvalues
out.res <- out.res %>% 
  mutate(qval = fdr) %>%
  mutate(nlq = -log10(qval)) %>%
  mutate(slq = sign(lfc) * nlq)

## remove redundant genes
idx <- tapply(seq(syms),syms,function(i){
  tmp <- out.res$nlq[i]
  max.i <- i[which.max(tmp)]
  return(max.i)
},simplify=T)

out.res <- out.res[idx,]
out.res <- out.res[-1,]

out.res <- DE_category(out.res,gs=mods)
names(out.res) <- c("logFC","P.value","-log10(P.value)","GENE_SYMBOL","adj.P.value","-log10(adj.P.value)","signed log adj.p","Gene_category")

```

#### Plots
```{r}
## one way to do this - go with ratio, not diff.
xl <- range(c(out.sen$logFC,out.res$logFC))
yl <- range(c(out.sen$`-log10(adj.P.value)`,out.res$`-log10(adj.P.value)`))

pdf(file=file.path(fig_dir1,"ACSOG_volcano_sen_all_smpls_v2.pdf"),width=7,height=7)
volcano_plot(out.sen) + ggtitle("Sensitive to AI (Pre vs Post) (180 pts)") + xlim(xl) + ylim(yl)
dev.off()

pdf(file=file.path(fig_dir1,"ACSOG_volcano_res_all_smpls_v2.pdf"),width=7,height=7)
volcano_plot(out.res) + ggtitle("Resistant to AI (Pre vs Post) (34 pts)") + xlim(xl) + ylim(yl)
dev.off()
```

## FGSEA - MSigDB 
### combine msigdb with mods
```{r}
msigdb <- readRDS(file.path(obj_dir,"Pathways","msigdb.rds"))

syms1 <- out.res$GENE_SYMBOL
mods1 <- lapply(mods,function(x){
  x <- unlist(x)
  x <- x[x %in% syms]
  return(x)
})
msigdb.1 <- c(msigdb,mods1)
```

### colors
```{r}
library(fgsea)
# bl <- colorRampPalette(c("white","steelblue"))(6)[5]
bl <- "steelblue"
mycols <- c(bl,brewer.pal(3,"Dark2"))
names(mycols) <- c("ERS","TNFa/NFkB","IFN-I","APM/TC")
```

### baseline
```{r}
sorted.bl <- out.bl %>% arrange(desc(`signed log adj.p`))
slps.bl <- sorted.bl[["signed log adj.p"]]
names(slps.bl) <- sorted.bl$GENE_SYMBOL
fg1.bl <- fgsea(pathways = msigdb.1, 
                  stats    = slps.bl,
                  minSize  = 15,
                  maxSize  = 500) %>%
  mutate(slp = -log10(padj) * sign(ES))

fg2.bl <- fg1.bl %>% 
  mutate(text = "") %>% 
  mutate(lab=1) %>%
  arrange(desc(slp)) %>%
  mutate(idx = seq(nrow(fg1.bl)))

fg2.bl$text[match(c(names(mods)),fg2.bl$pathway)] <- names(mods)

fg2.bl %>% filter(pathway %in% names(mods))

### violin plot of the FGSEA results highlighting slp and the custom modules
p <- ggplot(fg2.bl,aes(x=lab,y=slp)) + 
  geom_violin() +
  geom_point(data=fg2.bl %>% filter(!pathway %in% names(mods)),aes(y=slp),color="grey60",pch=95,size=4) +  
  geom_point(data=fg2.bl %>% filter(pathway %in% names(mods)),aes(y=slp,color=pathway),pch=95,size=10) +
  geom_text_repel(data=fg2.bl %>% filter(pathway %in% names(mods)),aes(label=text),size=3) +
  scale_color_manual(values=mycols) +
  theme_minimal() +
  theme(legend.position="none")

pdf(file=file.path(fig_dir1,"baseline_fgsea_violin.pdf"),width=3,height=7)
print(p)
dev.off()
```

### sensitive
```{r}
sorted.sen <- out.sen %>% arrange(desc(`signed log adj.p`))
slps.sen <- sorted.sen[["signed log adj.p"]]
names(slps.sen) <- sorted.sen$GENE_SYMBOL
fg1.sen <- fgsea(pathways = msigdb.1, 
                  stats    = slps.sen,
                  minSize  = 15,
                  maxSize  = 500) %>%
  mutate(slp = -log10(padj) * sign(ES))

fg2.sen <- fg1.sen %>% 
  mutate(text = "") %>% 
  mutate(lab=1) %>%
  arrange(desc(slp)) %>%
  mutate(idx = seq(nrow(fg1.sen)))

fg2.sen$text[match(c(names(mods)),fg2.sen$pathway)] <- names(mods)

fg2.sen %>% filter(pathway %in% names(mods))

### violin plot of the FGSEA results highlighting slp and the custom modules
p2 <- ggplot(fg2.sen,aes(x=lab,y=slp)) + 
  geom_violin() +
  ylim(c(-50,25)) +
  geom_point(data=fg2.sen %>% filter(pathway %in% names(mods)),aes(y=slp,color=pathway),pch=95,size=5) +
  geom_text_repel(data=fg2.sen %>% filter(pathway %in% names(mods)),aes(label=text),size=3) +
  scale_color_manual(values=mycols) +
  theme_minimal() +
  theme(legend.position="none")

pdf(file=file.path(fig_dir1,"sensitive_fgsea_violin.pdf"),width=4,height=7)
print(p2)
dev.off()
```

### resistance
```{r}
sorted.res <- out.res %>% arrange(desc(`signed log adj.p`))
slps.res <- sorted.res[["signed log adj.p"]]
names(slps.res) <- sorted.res$GENE_SYMBOL
fg1.res <- fgsea(pathways = msigdb.1, 
                  stats    = slps.res,
                  minSize  = 15,
                  maxSize  = 500) %>%
  mutate(slp = -log10(padj) * sign(ES))

fg2.res <- fg1.res %>% 
  mutate(text = "") %>% 
  mutate(lab=1) %>%
  arrange(desc(slp)) %>%
  mutate(idx = seq(nrow(fg1.res)))

fg2.res$text[match(c(names(mods)),fg2.res$pathway)] <- names(mods)

fg2.res %>% filter(pathway %in% names(mods))

### violin plot of the FGSEA results highlighting slp and the custom modules
p3 <- ggplot(fg2.res,aes(x=lab,y=slp)) + 
  geom_violin() +
  ylim(c(-50,25)) +
  geom_point(data=fg2.res %>% filter(pathway %in% names(mods)),aes(y=slp,color=pathway),pch=95,size=5) +
  geom_text_repel(data=fg2.res %>% filter(pathway %in% names(mods)),aes(label=text),size=3) +
  scale_color_manual(values=mycols) +
  theme_minimal() +
  theme(legend.position="none")

pdf(file=file.path(fig_dir1,"resistant_fgsea_violin.pdf"),width=4,height=7)
print(p3)
dev.off()
```

```{r}
fg2.s <- fg2.sen %>% select(pathway,slp)
fg2.r <- fg2.res %>% select(pathway,slp) 

fg2.sr <- fg2.r %>% 
  left_join(fg2.s,by="pathway") %>%
  mutate(text="")  

fg2.sr$text[match(c(names(mods)),fg2.sr$pathway)] <- names(mods)
fg2.sr$text[fg2.sr$slps.res < -10 & fg2.sr$slps.sen < -10] <- 
  fg2.sr$pathway[fg2.sr$slps.res < -10 & fg2.sr$slps.sen < -10]
names(fg2.sr) <- c("pathway","slps.res","slps.sen","text")

## asp =1

p4 <- ggplot(fg2.sr %>% filter(!text %in% names(mods)),   aes(x=slps.res,y=slps.sen)) +
  geom_hline(yintercept=0) +
  geom_vline(xintercept=0) +
  geom_point(color="grey80") +
  geom_point(data=fg2.sr %>% filter(text %in% names(mods)),aes(color=text)) +
  scale_color_manual(values=mycols) +
  geom_text_repel(data=fg2.sr %>% filter(text!=""),aes(label=text)) +
  # geom_abline(intercept = 0, slope = 1) +
  xlab("signed log adj.p (Resistant)") +
  ylab("signed log adj.p (Sensitive)") +
  ggtitle("FGSEA results (Resistant vs Sensitive)") +
  theme_minimal() +
  theme(legend.position="none")

pdf(file=file.path(fig_dir1,"changes_fgsea_sen_res.pdf"),width=7,height=7)
print(p4)
dev.off()
```
#### pathway analysis summary
```{r}
fg2.sr %>% filter(slps.res > 10)
fg2.sr %>% filter(slps.res < -10)

fg2.sr %>% filter(slps.sen > 10)
fg2.sr %>% filter(slps.sen < -10)
```

## Save analysis objects
```{r}
if(0){
  save(full_results,
       full_results_1,
       pd.need, 
       design,
       resis.design,
       sens.design,
       full_results_res,
       full_results_sens, 
       file=file.path(obj_dir,"ACOSOG","ACOSOG_data.rda"))
}else{
  x <-  load(file=file.path(obj_dir,"ACOSOG","ACOSOG_data.rda"))
}
```
