---
title: "2. Prepare TCGA data"
output: html_notebook
---

## Initial setup

### Load library
```{r}
library(dplyr)
```

### Set paths
```{r}
tcga_dir <- "~/Dropbox (HMS)/_projects_/B06 03 HR-APM/data/brca_tcga"
obj_dir <- "~/Dropbox (HMS)/_projects_/B06 03 HR-APM/data/RData"
```

## Load expresssion data (downloaded form cBioPortal)
```{r}
## expression matrix
ma.fns <- file.path(tcga_dir,
                    c("data_RNA_Seq_v2_expression_median.txt",
                    	"data_RNA_Seq_v2_mRNA_median_Zscores.txt",
                    	"data_RNA_Seq_v2_mRNA_median_all_sample_Zscores.txt"))


tcga.mat1 <- read.table(ma.fns[1],sep="\t",header=T)
tcga.mat2 <- read.table(ma.fns[2],sep="\t",header=T)
tcga.mat3 <- read.table(ma.fns[3],sep="\t",header=T)

tcga.mat1[1:5,1:5] # first two columns are Hugo Symbols and ENtrez Gene IDs
```

### Check column names
Column names of the three matrices are identical.
```{r}
identical(colnames(tcga.mat1),colnames(tcga.mat2)) # true
identical(colnames(tcga.mat1),colnames(tcga.mat3)) # true
```

## Pre-processing data
### Compare three expression matrices

```{r}
tcga.m1 <- as.matrix(tcga.mat1[,-(1:2)])
tcga.m2 <- as.matrix(tcga.mat2[,-(1:2)])
tcga.m3 <- as.matrix(tcga.mat3[,-(1:2)])

dim(tcga.m1) # 20531 genes, 1100 samples
dim(tcga.m2) # 20440 genes, 1100 samples
dim(tcga.m3) # 20531 genes, 1100 samples
```

### mat1 and mat3 have identical gene sets.
```{r}
identical(tcga.mat1$Hugo_Symbol,tcga.mat3$Hugo_Symbol) # true
```


### 7/1100 samples were duplicated (metastasized?)
```{r}
table(sub(".+\\.","",colnames(tcga.m3))) # 01 - primary, 06 - duplicated (metastasis?)

is.dup <- grepl("06$",colnames(tcga.m3))
sub("06","01",colnames(tcga.m3)[is.dup]) %in% colnames(tcga.m3) # TRUE
```

### Remove the duplicates
```{r}
tcga.m1 <- tcga.m1[,!is.dup]
tcga.m2 <- tcga.m2[,!is.dup]
tcga.m3 <- tcga.m3[,!is.dup]

colnames(tcga.m1) <- colnames(tcga.m2) <- colnames(tcga.m3) <- sub("\\.01$","",colnames(tcga.m3))
```


```{r}
### they are monotonously increasing relationships, so doesn't matter which samples to use for computing spearman correlation
syms12 <- intersect(tcga.mat1$Hugo_Symbol,tcga.mat2$Hugo_Symbol)
i=4
plot(tcga.m1[match(syms12[i],tcga.mat1$Hugo_Symbol),],
	tcga.m2[match(syms12[i],tcga.mat2$Hugo_Symbol),],
	main=syms12[i],
	xlab="M1",ylab="M2",	
	pch=20)
plot(tcga.m1[match(syms12[i],tcga.mat1$Hugo_Symbol),],
	tcga.m3[match(syms12[i],tcga.mat3$Hugo_Symbol),],
	main=syms12[i],
	xlab="M1",ylab="M3",
	pch=20)
plot(tcga.m2[match(syms12[i],tcga.mat2$Hugo_Symbol),],
  tcga.m3[match(syms12[i],tcga.mat3$Hugo_Symbol),],     
	main=syms12[i],	
	xlab="M2",ylab="M3",	
	pch=20)
```

## Load Clinical data
```{r}
cl <- readLines(file.path(tcga_dir,"data_bcr_clinical_data_patient.txt"))[-(1:4)]
clt <- do.call(rbind,strsplit(cl,split="\t"))
cn <- clt[1,]
tcga.clinical <- data.frame(clt[-1,],stringsAsFactors=F)
names(tcga.clinical) <- cn
tcga.clinical$PATIENT_ID <- gsub("-",".",tcga.clinical$PATIENT_ID)

used.ids <- colnames(tcga.m3)
if(0){
  table(used.ids %in% tcga.clinical$PATIENT_ID) ## all samples' records exist in clinical metadata
}

tcga.clinical <- tcga.clinical %>% filter(PATIENT_ID %in% used.ids)
```

## Disease status
```{r}
## # patients per disease
tcga.clin.1 <- tcga.clinical %>%
	filter(ER_STATUS_BY_IHC %in% c("Positive","Negative")) %>%
	filter(PR_STATUS_BY_IHC %in% c("Positive","Negative")) %>%
	filter(IHC_HER2 %in% c("Positive","Negative"))

mm <- cbind(
	`ER+`=(tcga.clin.1$ER_STATUS_BY_IHC=="Positive"),
	`PR+`=(tcga.clin.1$PR_STATUS_BY_IHC=="Positive"),
	`HER2+`=(tcga.clin.1$IHC_HER2=="Positive"))
rownames(mm) <- tcga.clin.1$PATIENT_ID

limma::vennDiagram(mm,main="TCGA samples")
```

### Patient IDs per category
```{r}
tcga.er.pos.ids <- (tcga.clinical %>% 
	filter((ER_STATUS_BY_IHC == "Positive" | PR_STATUS_BY_IHC == "Positive")
			& IHC_HER2 == "Negative"))$PATIENT_ID # 445

tcga.her2.pos.ids <- (tcga.clinical %>% 
	filter(ER_STATUS_BY_IHC == "Negative" & 
		PR_STATUS_BY_IHC == "Negative" &
		IHC_HER2 == "Positive"))$PATIENT_ID # 37

tcga.dp.ids <- (tcga.clinical %>% 
	filter((ER_STATUS_BY_IHC == "Positive" | PR_STATUS_BY_IHC == "Positive")& 
		IHC_HER2 == "Positive"))$PATIENT_ID # 126

tcga.tnbc.ids <- (tcga.clinical %>% 
	filter(ER_STATUS_BY_IHC == "Negative" & 
		PR_STATUS_BY_IHC == "Negative" &
		IHC_HER2 == "Negative"))$PATIENT_ID # 37

tcga.all.ids <- tcga.clinical$PATIENT_ID

tcga.ids <- list(
	hr=tcga.er.pos.ids,
	her2=tcga.her2.pos.ids,
	dp=tcga.dp.ids,
	tnbc=tcga.tnbc.ids,
	all = tcga.all.ids
)
sapply(tcga.ids,length)

## Be careful, Hugo symbols (or Entrez IDs) are not unique.
table(table(tcga.mat1$Hugo)) # not unique
table(table(tcga.mat2$Hugo)) # not unique

tcga.syms <- tcga.mat3$Hugo_Symbol ## genes whose expression is measured
```

## Save data in rda
```{r}
save(tcga.ids,tcga.clinical,tcga.syms,tcga.m1,tcga.m3,file=file.path(obj_dir,"tcga_RNA-seq.rda"))
```

